The influence of heavy metal stress on the evolutionary transition of teosinte to maize

Version 1

Abstract

Abstract

Maize originated from teosinte parviglumis following a subspeciation event occurred in volcanic regions of Mesoamerica. The elucidation of the phenotypic changes that gave rise to maize have focused on the direct consequences of domestication, with no insights on how environmental factors could have influenced specific gene function and human selection. We explored the impact of heavy metal (HM) stress by exposing both subspecies to sublethal concentrations of copper and cadmium. We also assessed the genetic diversity of loci encompassing three HM response genes affected by domestication: ZmHMA1, ZmHMA7 – encoding for heavy metal ATPases of the P1b family-, and ZmSKUs5, encoding a multicopper oxidase. ZmHMA1 and ZmSKUs5 map in the short arm of chromosome five, in a genomic region containing multiple linked QTLs with pleiotropic effects on domestication. A genomic analysis of the full chromosome shows that their loci show strong positive selection as compared to previously identified domestication genes. Exposure of teosinte parviglumis to HM stress results in a plant architecture reminiscent of extant maize, and upregulation of Teosinte branched1 (Tb1) in the meristem. ZmHMA1 and ZmHMA7 are expressed throughout development and respond to HM stress in both subspecies. ZmHMA1 is mainly involved in restricting plant height and optimizing the number of female inflorescences and seminal roots. Our results suggest that HM stress acted on specific ATPases involved in homeostasis, giving rise to phenotypic variants that were identified and selected by humans during domestication.

Introduction

Driven by either conscious or unconscious human selection, domestication can be considered a gradual evolutionary process resulting in the adaptation to agroecological environments and anthropogenic preferences (Ross-Ibarra et al. 2007). In the case of plants, the obvious consequences of domestication are represented by visually recognizable traits that include changes in photoperiod response, reduced capacity of seed dispersion, reduced lateral branching, larger fruits or grains, reduced duration of seed dormancy, and increase in the size of female inflorescences (Pickersgill 2007; Larson et al. 2014). Notably, research efforts to understand the phenotypic changes that gave rise to maize have mainly focused on the direct consequences of domestication in aerial organs and sometimes roots (Chen et al. 2022; Lopez-Valdivia et al. 2022; Ren et al. 2022), with no emphasis on the role of adaptive environmental responses that, by acting on specific gene function, could have given rise to phenotypic traits that were identified and selected by humans.

It is generally accepted that domestication of maize (Zea mays ssp. mays) from teosinte parviglumis (Zea mays ssp. parviglumis) initiated in Mesoamerica around 9,000 years before present (Matsuoka et al. 2002). The initial evolutionary transition that resulted in the event of subspeciation most likely occurred within the Balsas river drainage, in regions geographically converging with the Trans-Mexican Volcanic Belt, at the intersection of the States of Mexico, Guerrero, and Michoacan (Matsuoka et al. 2002). Although the nature of the initial phenotypic changes that allowed the emergence of primitive maize remains unclear, the most obvious architectural and morphological differences between both subspecies affect their vegetative and reproductive organs. Teosinte parviglumis generates numerous lateral branches ending in male inflorescences, while maize often has a few short or no lateral branches, not ending in male reproductive organs (Wills et al. 2013). Additionally, the seeds of teosinte parviglumis are covered by a hard glume that protects the kernel externally, while in maize glumes are reduced, soft and do not cover the kernel (Doebley and Stec 1993). Large scale efforts involving physical mapping and comparative genomic analysis have resulted in the identification of numerous regions showing differential nucleotide variability between teosinte parviglumis and the B73 reference maize genome. In some cases, mutations in genes located within those genomic regions result in phenotypes that reveal their involvement in domestication traits. One of the major genes affected by maize domestication is Teosinte branched1 (Tb1), a transcription factor of the TCP family that controls lateral branching, sexual determination, and glume formation (Doebley et al. 1997; Martín-Trillo and Cubas 2010). Tb1 acts as a repressor of lateral branching (Studer et al. 2017), giving rise to maize plants usually composed of a single stem. Other genes involved in domestication include Teosinte glume architecture1 (Tga1), ramosa2 (ra2) and Zea floricaula leafy2 (zfl2), among others (Bomblies and Doebley 2006; Sigmon and Vollbrecht 2010; Wang et al. 2015).

The importance of phenotypic plasticity is considered as likely relevant in the domestication history of crops (Ross-Ibarra et al. 2007; Lorant et al. 2020; Mueller et al., 2023); however, the role of environmental factors influencing their origins remains elusive and poorly investigated. A previous study showed that teosinte parviglumis grown under low temperatures and CO2 concentrations - reminiscent of those prevailing in the late Pleistocene and early Holocene - exhibit some phenotypes related to the domestication syndrome, including short branches, a single main stalk, and partially open seed fruitcases (Piperno et al. 2015; Piperno et al. 2019); however, the genetic basis of these changes have not been investigated. Although these results suggest that phenotypic plasticity could have influenced the domestication process, to this date there are no hints on how the adaptation to local environmental conditions could have affected specific gene function and guided human selection.

Recent genomic evidence indicates that admixture of maize with native teosinte mexicana populations (Zea mays ssp. mexicana) contributed to maize adaptation across a wide range of altitudes and latitudes in the American continent (Yang et al. 2023); however, the earliest maize macro-specimens found to date lack teosinte mexicana admixture and exhibit morphological attributes of fully domesticated cobs, indicating that mexicana is not a direct ancestor of maize and that its introgression occurred after initial divergence from teosinte parviglumis (Vallebueno-Estrada et al. 2023; Acosta-Bayona et al. 2024). Consequently, the nature of the genetic factors that guided the emergence of maize remain largely unknown, and the initial phenotypic changes that caused the original divergence of these two subspecies have not been identified. It is also unknown if domestication initiated by direct action of human selection on teosinte natural standing variation, or if the response to specific environmental factors generated phenotypic changes that were identified and selected by humans to act on native teosinte parviglumis populations.

Maize can adapt to abiotic factors such as heavy metal (HM) stress through numerous mechanisms that prevail during development (AbdElgawad et al. 2020; Gallo-Franco et al. 2020; Malenica et al. 2021; Gao et al. 2022). These include metal immobilization by mycorrhizal association, metal sequestration, or complexation by exuding organic compounds from roots, metal intracellular sequestration and compartmentalization in vacuoles. HMs include essential elements that are necessary for maize growth through enzymatic function, such as iron (Fe), manganese (Mn), zinc (Zn), magnesium (Mg), molybdenum (Mo), and copper (Cu); and non-essential elements like Cadmium (Cd), chromium (Cr), lead (Pb), aluminum (Al), and selenium (Se). In general, HM stress leads to visibly reduced plant growth due to the reduced cell elongation and cell wall elasticity. At the molecular level, metal toxicity can disturb the cellular redox balance and lead directly or indirectly to oxidative stress and the formation of reactive oxygen species (ROS), shifting the redox balance to the pro-oxidative side until the antioxidative defense system allows a new balanced redox status (Smeets et al. 2009; Hall 2002).

The comparison of the Palomero toluqueño and B73 genomes resulted in the identification of a large collection of identical sequence regions (IdSRs) with low nucleotide variability and transcriptional evidence of expression (Vielle-Calzada et al. 2009). Putative domestication loci included ZmHMA1 and ZmHMA7, two genes encoding members of the P1b-ATPases protein superfamily, also known as Heavy Metal ATPases (P-type ATPases) characterized by the generation of a phosphorylated intermediate during the transport reaction cycle. Members of this superfamily contain 8 to 12 transmembrane domains as well as conserved ATP binding, phosphorylation and dephosphorylation sites (Solioz and Vulpe 1996; Axelsen and Palmgren 2001; Voskoboinik et al. 2003; Williams and Mills 2005). The superfamily can be divided in subfamilies that include H±ATPases (type 3A) in plants and fungi, Na+/K ATPases (type 2C/D) in animals, Ca2±ATPases (type 2A/B), and heavy metal (HM) transporting ATPases (type 1B) capable of transporting a large variety of HM cations across cell membranes. ZmHMA1 and ZmHMA7 are type 1B ATPases that include a conserved Cys-Pro-Cys/His/Ser motif within a transmembrane domain necessary for the translocation of HMs cations through the membrane channel. A third Palomero toluqueño/B73 IdSR included Skewed Root Growth Similar5 (ZmSKUs5), a gene encoding a Skewed5 (SKU5)-similar (SKS) protein classified as a multi-copper oxidase containing cuperdoxin domains (Sedbrook et al. 2002). All three genes map to chromosome five (chr.5), with ZmHMA1 and ZmSKUs5 within a genomic region that fractionates into multiple QTLs under selection during the parviglumis to maize transition (Doebley and Stec 1993; Lemmon and Doebley, 2014). Contrary to major domestication loci showing a single highly pleiotropic gene responsible for important domestication traits, in this chr.5 genomic region the smallest 1.5-LOD support interval contained more than 50 genes, suggesting that phenotypic effects are due to multiple linked QTLs (Lemmon and Doebley, 2014). While the functional characterization of all three maize genes remains elusive, the Arabidopsis and barley ZmHMA1 homologs act during heavy metal transport in the chloroplast (Boutigny et al. 2014), demonstrating their functional involvement in homeostasis.

To investigate a possible involvement of heavy metal response in the origin and domestication of maize, we compared the phenotypic changes that teosinte parviglumis and maize undergo when grown in the presence of sublethal concentrations of Cu and Cd. We show that teosinte parviglumis responds to HM stress by acquiring an aerial architecture reminiscent of extant maize, and that Tb1 expression is upregulated in teosinte parviglumis meristems when grown under HM stress. A large-scale genomic analysis of nucleotide variability across maize chr.5 indicates that loci encompassing ZmHMA1, ZmHMA7, and ZmSKUs5 show reduced nucleotide variability in maize as compared to teosinte parviglumis. Both ZmHMA1 and ZmHMA7 are active throughout maize and teosinte parviglumis development, and their expression responds to HM stress. In addition, mutations in ZmHMA1 show phenotypic defects related to the maize domestication syndrome, some of which are only manifested under exposure to HMs. Overall, our results indicate that the evolutionary transition that gave rise to extant maize was influenced by heavy metal stress through the activity of P1b-ATPases.

Results

Teosinte parviglumis responds to heavy metal stress by acquiring an architecture reminiscent of extant maize

To compare HM response in teosinte parviglumis and maize, we exposed wild-type W22 maize inbred and CIMMYTMA 29791 teosinte parviglumis individuals to sublethal concentrations of Cu (400 mg/kg of CuSO4) and Cd (16mg/kg of CdCl2). CuSO4 and CdCl2 were homogenized in soil before transplant and grown under controlled greenhouse conditions. HM concentrations were based on preliminary dose responses that allowed maize plants to reach reproductive stages and complete their life cycle (AbdElgawad et al. 2020). As previously reported (Atta et al. 2023), maize individuals grown in the presence of HMs showed a significant reduction in aerial traits such as stem diameter, size of leaves, size and number of roots, and size of female inflorescences (Figure 1; Tables 2 and S2). On average, stem diameter showed a reduction of about 17% as compared to plants grown in absence of HM stress (Figure 1f to 1j). Similarly, the length and width of developing leaves showed average reductions of 12%, and mature leaves of 6%. Reproductive traits such as the number of female inflorescences (NFI) and their average weight (WFI) showed a significant reduction (NFI = 1.20 ± 0.42 in absence of HMs and NFI = 0.70 ± 0.67 in presence of HMs; WFI = 40.06 ± 14.39 in absence of HMs and WFI = 37.75 ± 18.83 in presence of HMs; Table 2). As expected, the most significant effects occurred in the root system, with a significant increase in the number of brace roots under HMs (NBR = 12.90 ± 4.33 in absence of HMs; NBR = 16.30 ± 4.99 in presence of HMs; Table 2), and a significant decrease in their length (LBR = 16.15 ± 9.75 cm in absence of HMs; LBR = 4.00 ± 2.04 cm in presence of HMs; Table 2). The number and length of crown roots were reduced on average by 33% and 18%, respectively (NCR = 11.30 ± 2.11 in absence of HMs; NCR = 7.60 ± 0.97 in presence of HMs; LCR = 58.90 ± 21.87 cm in absence of HMs; LCR = 48.85 ± 11.80 cm in presence of HMs; Table 2). Nodal roots also showed a significant reduction in their number and length (NNR = 10.50 ± 2.27 in absence of HMs; NNR = 6.70 ± 2.79 in presence of HMs; LNR = 87.20 ± 16.80 cm in absence of HMs; LNR = 55.30 ± 21.10 cm in presence of HMs; Tables 2 and S2). By contrast, the number (NSR) and length (LSR) of seminal roots was significantly increased in the presence of HMs (NSR = 4.10 ± 1.45 in absence of HMs; NSR = 6.60 ± 1.35 in presence of HMs; LSR = 27.70 ± 13.38 cm in absence of HMs; LSR = 43.50 ± 14.01 cm in presence of HMs) when measured 50 days after transplant (Table 2).

Phenotypic response of teosinte parviglumis and wild-type W22 maize grown under heavy metal stress.

(a) to (e), teosinte parviglumis; (f) to (j), W22 maize. (a) Teosinte parviglumis in absence of heavy metal stress at early flowering stage. (b) Teosinte parviglumis grown under heavy metal stress at early flowering stage. (c) Teosinte parviglumis in absence of heavy metal stress at late flowering stage. (d) Teosinte parviglumis grown under heavy metal stress at late flowering stage. (e) Teosinte parviglumis showing proliferation of female inflorescences arising from independent axillary meristems under heavy metal stress (detail of d). (f) W22 maize in absence of heavy metal stress at early flowering stage. (g) W22 maize grown under heavy metal stress at early flowering stage. (h) W22 maize in absence of heavy metal stress at late flowering stage. (i) W22 maize grown under heavy metal stress at late flowering stage. (j) W22 maize female inflorescences at late flowering stage (detail of i). Scale bars: 10 cm.

Interestingly, teosinte parviglumis exhibited a maize-like phenotype when grown in the presence of HMs, with complete absence of tillering in most plants (Figure 1; Tables 1, S1 and S3). From a total of 30 plants tested, 28 showed total absence of tillering, and two initiated secondary stems during vegetative stages but arrested their growth early during development, resulting in the presence of a single dominant stem at reproductive stages (Figure 1a to 1d). Additionally, all 30 plants tested showed complete absence or residual lateral branching, with unusual proliferation of female inflorescences arising from independent axillary meristems in primary stems (Figure 1e). This significant increase in the number of female inflorescences clustered within a single stem (NFI = 1.80 ± 0.63 in absence of HMs; NFI = 3.90 ± 1.66 in presence of HMs; Figure 1e and Table 1) was associated with shorter flowering time as compared with normal growing conditions in absence of HMs (Figure 1d). In addition, teosinte parviglumis grown in the presence of HMs showed significantly reduced leaf length (LIL = 71.30 ± 11.32 cm in absence of HMs; 56.25 ± 7.80 cm in presence of HMs; Table 1). The absence of lateral branches resulted in lower values of aerial dry weight (DWAP = 133.49 ± 28.28 gr in the absence of HMs; DWAP = 82.72 ± 11.67 gr in presence of HMs; Table 1). Teosinte parviglumis grown under HMs showed a significant reduction in the length of brace roots (LBR = 20.00 ± 11.49 cm in absence of HMs; LBR = 2.88 ± 0.66 cm in presence of HMs; Table 1), but no significant changes in their number. The number and length of crown and nodal roots was not affected by HM stress, nor did the presence of HMs modify the absence of seminal roots intrinsic to teosinte parviglumis accessions (Perkins and Lynch 2021). Taken together, these results suggest that the response of maize and teosinte parviglumis to HM stress can give rise to phenotypes previously shown to distinguish both subspecies at the light of the domestication syndrome.

Genomic regions encompassing ZmHMA1, ZmHMA7 and ZmSKUs5 show reduced nucleotide variability in extant maize as compared to teosinte parviglumis

ZmHMA1 and ZmSKU5 mapped 52.1 and 41.8 cM, respectively, in a genetic map of chr.5 calculated to be 86.64 cM, with an average megabase pair to centimorgan ratio of 1.873 Mbp/cM (Lemmon and Doebley, 2014; Figure S1). The corresponding mapping population used isogenic recombinant inbred lines (NIRILs) with informative cross-over events concentrated in the region of interest and replicated phenotypic assessments in block experiments. Multiple QTLs were detected in a region spanning across 82 cM, none with single large effect genes such as those found in other chromosomes (Wang et al., 2005; Studer et al. 2011; Wills et al., 2013), suggesting that several associated domestication traits are due to multiple linked QTLs of small effect (Lemmon and Doebley, 2014). The location of ZmSKU5 overlaps with a single wide 1.5-LOD support interval that spreads across 50.6 cM (number of kernels per rank; Figure S1). By contrast, ZmHMA1 overlaps with a contiguous region containing no less than five QTL 1.5- LOD support intervals (ear diameter; number of kernels per rank; percentage of staminate spikelets; tassel branch number; and tillering index; Figure S1).

To determine if the regulatory and coding regions of ZmHMA1, ZmHMA7, and ZmSKUs5 were affected by human selection, we calculated their level of nucleotide variability based on single nucleotide polymorphisms (SNPs) for all extant maize and teosinte parviglumis accessions included in the diversity panel HapMap3, using the corresponding sequences of Tripsacum dactyloides as an outgroup. The levels of the nucleotide variability index - π(m) and π(t) for maize and teosinte parviglumis, respectively - were determined in bins of 100 nt using VCFtools, including alignments encompassing a genomic region that included 15 kb upstream of the START codon, the corresponding coding region for each gene, and 15 kb downstream of the STOP codon (Figure 2 and Table S4). In the case of ZmHMA1, the upstream region values were π(m)= 0.00071 and π(t)= 0.00870, and for the downstream region, π(m)=0.00097 and π(t)=0.00248. By contrast, the ZmHMA1 coding sequence showed low values in both subspecies: π(m)= 0.0050 and π(t)= 0.00106. These results indicate that human selection in the ZmHMA1 locus preferentially affected its upstream and downstream regulatory regions (Figure 2a and Table S4). Interestingly, although maize π values are at least one order of magnitude lower than π values in teosinte, π(t) values are also one order of magnitude lower than π values of Tripsacum sp., suggesting that a genetic bottleneck also affected teosinte parviglumis in these loci. ZmHMA7 showed contrasting values of π(m) and π(t) across the entire coding sequence and its regulatory regions (Figure 2b). For the upstream region, values were π(m)= 0.0009 and π(t)=0.00411; for the ZmHMA7 coding region, π(m)= 0.00015 and π(t)= 0.00148; and for the downstream region, π(m)=0.00066 and π(t)=0.00454. ZmSKUs5 also showed contrasting nucleotide variability in the entire coding and flanking regions (Figure 2c and Table S4). Values were π(m)= 0.0011 and π(t)= 0.00735 for the upstream region, π(m)= 0.00044 and π(t)= 0.00754 for the coding sequence, and π(m)=0.00203 and π(t)=0.00626 for the downstream region (Table S4). For all three genes, similar π values in the coding region of maize landraces and improved lines indicates that the loss of nucleotide variability occurred during the teosinte parviglumis to maize transition, and not during subsequent contemporary breeding (Table S4). By contrast, the ZmGLB1 neutral control locus shows equivalent levels of nucleotide variability in maize and teosinte parviglumis (Eyre-Walker et al. 1998).

Genetic diversity of loci encompassing ZmHMA1, ZmHMA7, and ZmSKUs5.

The nucleotide variability index (π) was calculated for all HapMap3 accessions of teosinte parviglumis (blue line), maize (red line), and Tripsacum dactyloides (orange line), taking in consideration the coding sequence (arrow) and an upstream and downstream region of 15 Kb encompassing each gene. (a) Nucleotide variability in the ZmHMA1 locus. (b) Nucleotide variability in the ZmHMA7 locus. (c) Nucleotide variability in the ZmSKUs5 locus. (d) Nucleotide variability in the ZmGLB1 locus.

We also compared the level of genetic variability in ZmHMA1, ZmSKUs5, and ZmHMA7 (and their neighboring regions) to all chr.5 genes categorized as likely affected by domestication (Hufford et al., 2012), and to the full collection of neutral genes distributed across the whole chromosome. The analysis was conducted using 100 bp bins and sliding windows of 50 bp; for a total of 385,829 bins in the case of chr.5 domestication genes, and 2,373,010 bins for neutral genes. As shown in Figure S2, the levels of nucleotide variability in the three candidate genes and their six neighboring loci are significantly reduced in maize landraces as compared to teosinte parviglumis inbred lines. When compared to the full neutral gene set of chr.5, this reduction is more severe than the reduction affecting 304 previously identified domestication genes, confirming the exceptional selective pressure that was imposed in those three loci during the evolutionary transition from teosinte parviglumis to extant maize.

ZmHMA1 and ZmHMA7 are active throughout development and their expression is influenced by the presence of heavy metals

The gene structure of ZmHMA1, ZmHMA7, and ZmSKUs5 is shown in Figure 3. The coding sequence of ZmHMA1 and ZmHMA7 is approximately 21.6 Kb and 4.5 Kb, respectively, while ZmSKUs5 is approximately 4.2 Kb in length. Within the short arm of chromosome 5, ZmHMA1 and ZmSKUs5 are separated by 9.16 Mb, with ZmHMA1 approximately 15 Mb away from the maximum likelihood prediction of a QTL involved in maize domestication (Doebley and Stec 1993; Vielle-Calzada et al. 2009). ZmHMA7 maps to the centromeric region of chromosome 5. A detailed PlantCARE-based analysis identified several conserved binding sites and regulatory elements in the promoter region of these three genes (Lescot et al. 2002). ZmHMA1 and ZmHMA7 share binding sites to basic Helix-Loop-Helix (bHLH) transcription factors, as well as abscisic acid response (ABRE), ethylene responsive (ERF), and G-box (CACGTG) elements recognized by leucine zipper family (bZIP) and bHLH transcription factors (Figure 3a). The regulatory region of both ZmHMA1 and ZmSKUs5 contain a binding site for MIKC-type MADS box transcription factors (Figure 3a), whereas ZmHMA7 and ZmSKUs5 share a jasmonic acid-response TGACG-motif (Figure 3a), as well as a Myb-binding (MBS) and GATA binding site. In addition, ZmHMA1 contains a binding site for TCP proteins such as Tb1, increasing the number of families of transcription factors that could be involved in shared or individual regulation of these three genes (Figure 3a).

Structure and gene expression of ZmHMA1 and ZmHMA7 in teosinte parviglumis and W22 maize.

(a) Structural analysis of ZmHMA1, ZmHMA7, and ZMSKUs5. Blue triangles indicate the location of motifs predicted in the regulatory region of each gene. The insertion site of Mu line LANMu1029790 is indicated by a red triangle. (b) RT-PCR analysis of ZmHMA1 and ZMHMA7 expression throughout development of teosinte parviglumis and W22 maize in presence or absence of heavy metal stress; L=leaf; S=stem; PR=primary root; NR=nodal root; CR=crown root. V2: 15 days after transplant; V6: 30 days after transplant; R1: 50 days after transplant. Controls: (-) absence of DNA; (+) genomic DNA. (c) qPCR analysis of ZmHMA1 expression in 1-month old primary roots, in absence or presence of heavy metal stress. (d) qPCR analysis of ZmHMA7 expression in 1-month old primary roots, in absence or presence of heavy metal stress. * P<0.01; ** P<0.001.

To elucidate the expression patterns of ZmHMA1 and ZmHMA7, we conducted reverse transcriptase PCR (RT-PCR) in leaves, stems, roots, and flowers at two vegetative (V2 and V6) and one reproductive (R1) developmental stage, in teosinte parviglumis and wild-type W22 maize individuals (Figure 3b). In the case of maize, ZmHMA1 was expressed in leaves, stems, roots and flowers throughout development in either absence or presence of HMs. ZmHMA7 was also expressed in aerial organs throughout development in both absence and presence of HMs but was not expressed in roots at reproductive stage R1. In teosinte parviglumis, ZmHMA1 was also expressed throughout development in either absence or presence of HMs. By contrast, ZmHMA7 was constitutively expressed in aerial organs, but only expressed in roots under HM stress during V2 and V6 stages. To further determine if the level of expression of these two genes is influenced by the presence of HMs, we conducted quantitative real-time PCR (qPCR) using the primary root of teosinte parviglumis and maize plantlets at four weeks after germination (Figures 3c and 3d). In wild-type W22 maize, the expression of ZmHMA1 was not significantly different in the presence and absence of HMs. By contrast, in teosinte parviglumis the expression of ZMHMA1 was significantly increased in presence of HMs (Figure 3c). In the case of ZmHMA7, expression was significantly increased in both maize and teosinte parviglumis when plants were grown in the presence of HMs (Figure 3d). Overall, these results indicate that the activity of both ZmHMA1 and ZmHMA7 is influenced by HM stress in teosinte parviglumis and maize.

ZmHMA1 null mutant individuals show phenotypic defects related to the domestication syndrome

Transposon Mu insertional lines have been widely used to elucidate gene function in maize. To elucidate the function of ZmHMA1, we phenotypically analyzed LANMu1029790, a 4- generation backcross of line mu1029790 into the W22 background selected for single copy insertions in ZmHMA1 (McCarty et al. 2005). In LANMu1029790, a single Mu insertion is in the first exon of ZmHMA1, 65 bp downstream of the START codon (Figure 3a). RT- PCR analysis showed that expression of ZmHMA1 is completely absent in homozygous individuals for the LANMu1029790 insertion (Figure S3), confirming that the corresponding mutant is a null allele. Plants harboring a homozygous LANMu1029790 insertion are subsequently designated as zmhma1-m1::MuDR (abbreviated zmhma1-m1).

To determine the function of ZmHMA1 in maize, we compared the phenotype of wild-type W22 and homozygous zmhma1-m1 individuals at the reproductive R1 stage (Table 2). Under normal growing conditions, homozygous zmhma1-m1 individuals generated significantly less leaves than wild-type (NL = 13.4 ± 0.84 in wild-type, and NL = 11.4 ± 0.84 in zmhma1-m1; P<0.001). Whereas the length of mutant leaves was smaller than wild-type (LML = 77.4 ± 6.64 cm in wild-type, and LML = 69.85 ± 9.18 cm in zmhma1-m1), their width was significantly larger (WML = 6.75 ± 1.3 cm in wild-type, and WML = 9.75 ± 1.4 cm in zmhma1-m1; P<0.001), resulting in an increase of the total dry weight of aerial organs in mutant plants (DWAP = 18.43 ± 8.04 grams in wild-type, and DWAP = 28.46 ± 9.54 grams in zmhma1-m1). Homozygous zmhma1-m1 individuals also produced significantly more female inflorescences per plant (NFI = 1.2 ± 0.42 in wild-type, and NFI = 1.9 ± 0.32 in zmhma1; P<0.001) a difference that reflected in the total dry weight of female inflorescences per plant (WFI = 40.06 ± 14.39 gr in wild-type and WFI = 69.5 ± 11.2 gr in zmhma1; P<0.001); and significantly more spikes per male inflorescence (NSB = 9 ± 1.63 in wild-type, and NSB = 13.4 ± 2.32 in zmhma1-m1; P<0.001). Interestingly, the general length of roots was significantly smaller in homozygous zmhma1-m1 individuals, in particular nodal roots (LNR = 87.2 ± 16.8 cm in wild-type, and LNR = 50.9 ± 12.83 cm in zmhma1-m1; P<0.001); however, mutant plants showed a significant increase in the number of seminal roots per plant (NSR = 4.1 ± 1.45 in wild-type, and NSR = 7.3 ± 1.16 in zmhma1-m1; P<0.001). When wild-type and mutant individuals were grown under HM stress, differences in the size of leaves, number of female inflorescences per plant, and number of spikes per male inflorescence were maintained. In addition, mutant individuals exposed to HM stress were significantly taller than wild-type (HEI = 82.2 ± 34.94 cm in wild-type, and HEI = 125.9 ± 18.81 cm in zmhma1-m1; P<0.01). In addition, total root and seminal root length were equivalent in wild-type and mutant individuals, mainly because the length of wild-type roots was significantly reduced under HM stress; however, the number of seminal roots was again significantly higher in mutant individuals (NSR = 6.6 ± 1.35 in wild-type, and NSR = 8.2 ± 1.62 in zmhma1-m1; P<0.05). These results indicate that in maize aerial organs, ZmHMA1 is involved in promoting leaf growth, restricting the number of female inflorescences per plant, and restricting the number of spikes in male inflorescences. In underground organs, ZmHMA1 is involved in promoting root growth and restricting the number of seminal roots per individual. Under HM stress, ZmHMA1 shows an additional role in restricting plant height, a function that was not identified under normal growing conditions. Overall, these results suggest that ZmHMA1 is involved in the control of some of the aerial and underground phenotypic traits that distinguish maize from teosinte parviglumis, especially under HM stress conditions.

Under heavy metal stress, Tb1 is overexpressed in the apical meristem of teosinte parviglumis

The absence of secondary ramifications in teosinte parviglumis grown under HM stress is reminiscent of the aerial architecture of wild-type maize. Previous QTL and molecular analysis suggested that Tb1 contributed to the architectural difference between maize and teosinte, as tb1 maize mutants develop secondary stem ramifications ending in male inflorescences, resembling teosinte in their overall architecture (Doebley et al. 1997) Interestingly, the regulatory region of ZmHMA1 contains a specific binding site for members of the Teosinte-branched 1/Cycloidea/Proliferating (TCP) family of transcription factors that includes Tb1 (Figure 3a). To investigate the possibility that the phenotypic response of teosinte parviglumis under HM stress could be regulated by Tb1, we conducted quantitative real-time PCR (qPCR) to determine the expression of ZmHMA1 and Tb1 in the shoot apical meristem (SAM) of one month old W22 maize and teosinte parviglumis plantlets (Figure 4a and 4b). The expression ZmHMA1 was downregulated in W22 wild-type maize meristems in the presence of HMs. By contrast, HM stress resulted in a significant increase of ZmHMA1 expression in teosinte parviglumis, confirming that in the maize ancestor the level of ZmHMA1 activity is influenced by the presence of HMs in the soil. In addition, the expression of Tb1 was significantly upregulated in teosinte parviglumis meristems in the presence of HMs, suggesting that formation of secondary ramifications in teosinte parviglumis is repressed by Tb1 function under HM stress. Taken together, these results open the possibility for a direct positive regulation of ZmHMA1 by transcriptions factor of the TCP family, including Tb1.

Tb1 and ZmHMA1 expression in the shoot apical meristem of plantlets grown under heavy metal stress.

(a) qPCR analysis of Tb1 expression in 1-month old teosinte parviglumis plantlets grown in absence or presence of heavy metal stress. (b) qPCR analysis of ZmHMA1 expression in 1-month old teosinte parviglumis and W22 maize plantlets in absence or presence of heavy metal stress. * P<0.01.

Discussion

Although the evolutionary transition from teosinte parviglumis to maize resulted in important phenotypic divergence between landraces and their wild progenitor, the relevance of teosinte parviglumis phenotypic plasticity and its genetic consequences during domestication remain poorly understood. Our study offers the first evidence showing that a specific environmental factor (heavy metal stress) influenced specific gene function (heavy metal response ATPases) and human selection during the parviglumis to maize transition. The comparison of HM responses between both subspecies indicates that teosinte parviglumis phenotypically responds to HM stress acquiring phenotypes reminiscent of maize. Whereas low temperatures and CO2 concentrations also caused a significant decrease in lateral branching, these environmental response has not been related to specific gene function and human selection. The absence of tillering exhibited by teosinte parviglumis under HM stress was not reported under conditions simulating an early Holocene environment (Piperno et al., 2015). Our results also show an increase in the number of female inflorescences per plant. This HM stress-dependent proliferation of several closely associated female inflorescences in a single teosinte parviglumis stem has not been previously reported. Female inflorescences arose from individual axillary meristems in multiple internodes, significantly increasing teosinte parviglumis seed yield. Our evidence indicates that HM stress revealed a teosinte parviglumis environmental plasticity that is directly related to the function of specific HM response genes that were affected by domestication through human selection.

We also show that ZmHMA1, ZmHMA7 and ZmSKUs5 exhibit reduced genetic diversity in extant maize as compared to teosinte parviglumis, suggesting that all three loci were affected by the evolutionary transition that gave rise to maize. The maize regulatory region and downstream 3’ area of the ZmHMA1 locus exhibit pronounced reduction in SNPs compared to teosinte parviglumis lines, likely affecting the molecular scope of its regulatory proteins, including transcription factors of the TCP family (Martín-Trillo and Cubas 2010). By contrast, ZmHMA7 and ZmSKUs5 showed a selective sweep across the entire maize locus. Although loss of genetic diversity is usually the result of human selection during domestication, it can also represent a consequence of natural selective pressures favoring fitness of specific teosinte parviglumis allelic variants better adapted to environmental changes and subsequently affected by human selection during the domestication process. The significant reduction of nucleotide variability in the ZmHMA1 locus of teosinte parviglumis suggests that indeed that a natural selection pressure Examples of edaphological factors driving genetic divergence in the teosintes or maize include local adaptation to phosphorus concentration in mexicana and parviglumis (Aguirre-Liguori et al. 2019), and fast maize adaptation to changing iron availability through the action of genes involved in its mobilization, uptake, and transport (Benke and Stich 2011).

The maize genome contains 12 genes encoding for members of the P1b family of ATPases (Cao et al. 2019). They are subdivided according to their preferential substrate affinity: the Copper (Cu)/Silver (Ag) group includes ZmHMA1 to ZmHMA4; and the Zinc (Zn)/Cobalt (Co)/Cadmium (Cd)/Lead (Pb) includes ZmHMA5 to ZmHMA12. Initial studies suggested that ZmHMA1 is expressed at low levels in vegetative, reproductive or root organs throughout development, contrary to ZmHMA7 that is strongly expressed in developing seeds 10 days after pollination; however, both genes appeared downregulated in plants grown under either Cu or Cd stress (Zhiguo et al. 2018). Additional studies by Cao et al (2019) indicated low expression of ZmHMA1 and high levels of ZmHMA7 expression in 12 tissues: roots, nodes, shoots, leaves, tassel, shoot apical meristem and young stem, internodes, anthers, silks, whole seedlings, endosperm, embryo, and pericarp. Under Cd stress, expression of both ZmHMA1 and ZmHMA7 was significantly upregulated in roots and downregulated in stems (Cao et al. 2019), contradicting previous results by Zhiguo et al. (2018). In Arabidopsis and barley, the corresponding ZmHMA1 structural homologs (AtHMA1 and HvHMA1, respectively) can transport a wide range of metals into the chloroplast (Mikkelsen et al. 2012; Boutigny et al. 2014). In the case of the Arabidopsis structural homolog of ZmHMA7, AtHMA7 (also called RAN1) is proposed to act as a metallochaperone located in the Golgi apparatus, delivering Cu to the ethylene receptor located in the plasma membrane (Hirayama et al. 1999). Our evidence complements previous studies by showing that, in maize, ZmHMA1 root expression is restricted to vegetative developmental stages, and that ZmHMA7 is expressed in roots in response to HM stress. We also show through qPCR analysis that, under HM stress, the expression of both ZmHMA1 and ZmHMA7 is upregulated in teosinte parviglumis primary roots of 4- week-old plantlets. Overall, these results confirm that both ZmHMA1 and ZmHMA7 are heavy metal response genes active throughout teosinte and maize development.

In maize, the phenotypic analysis of null mutant individuals indicates that ZmHMA1 is involved in the control of aerial and underground developmental traits, some of which are related to important functional adaptations that distinguish maize from its wild ancestor. In addition to changes in leaf number, leaf length, and number of spikes per male inflorescence, zmhma1-m1 mutant individuals grown under normal edaphological conditions showed a significant increase in the number of female inflorescences per plant, a significant reduction in the length of nodal roots, and a significant increase in the number of seminal roots per plant. Whereas the promotion of nodal root growth is a distinctive trait of maize phenotypic plasticity and adaptation, previous studies showed that contrary to maize, plants of teosinte parviglumis lack seminal roots. This absence is also characteristic of paleobotanical maize specimens dating more than 5,000 years before present, confirming that the acquisition of seminal roots is an important trait acquired during domestication (Lopez-Valdivia et al. 2022). The presence of seminal roots distinguishes extant maize from the teosintes (Burton et al. 2013), suggesting a possible involvement in domestication (Perkins and Lynch 2021; Lopez-Valdivia et al. 2022). Our results suggest that the allelic variants of the ZmHMA1 locus selected during domestication could have contributed to promote seminal root formation in maize, optimizing their number in accordance to seed endosperm availability and carbohydrate reserves (Perkins and Lynch 2021; Lopez-Valdivia et al. 2022). Interestingly, ZmHMA1 plays an important role in contributing to restricting plant height in response to HM stress, a phenotype that is also an important distinction between teosinte parviglumis and maize. Taken together, these results indicate that ZmHMA1 function was involved in the phenotypic transition from teosinte parviglumis to maize through the influence of HM stress; however, a better definition of its role in the origin and domestication of maize will require a functional analysis of the ZmHMA1 teosinte parviglumis homolog, and a detailed molecular analysis of the corresponding allelic variants and their phenotypic effects in both subspecies.

Ideas and Speculation

Our results point to an evolutionary model in which environmental changes in the form of edaphological HM stress imposed natural selective forces that acted on native teosinte parviglumis populations, causing phenotypic responses and a severe genetic bottleneck (Figure S4). Adapted plants developed phenotypic traits such as shorter height and emergence of seminal roots, progressively giving raise to primitive maize individuals with reduced genetic variation in HM response genes such as ZmHMA1. The subsequent action of human selection caused a drastic reduction in allelic variation of heavy metal response genes, and genetic fixation of the resulting domestication traits.

The most important source of HMs in ancient soils of Mesoamerica is volcanic activity through short- and long-term effects related to ores, hydrothermal water flow and ash deposits. The history of eruptive activity in the Trans Mexican Volcanic Belt during the early Holocene coincides with the temporal estimation of initial stages in the evolutionary transition that gave rise to maize. Intense and recurrent eruptions occurred 24,000 to 6,000 years ago near the region of the Balsas River basin presumed to be the cradle of maize domestication (Matsuoka et al. 2002; Macias 2005). Located less than 100 Km from the region that harbors the natural teosinte parviglumis populations most closely related to maize (Matsuoka et al. 2002), the Nevado de Toluca volcano underwent four major eruptions approximately 24100, 12100, 10500, and 8500 years before present (yr BP; Macias 2005), causing in the region a period of climatic instability that lasted until 8500 yr BP (Ludlow-Wiechers et al., 2005). Could volcanic eruptions during the early Holocene have modified soil composition and influence the initial evolutionary transition from teosinte parviglumis to primitive maize? The possibility of environmental cues influencing maize emergence was mentioned by Benz and Iltis (1992), but not in relation to volcanic activity. Deciphering the ancient DNA sequence of HM response genes such as ZmHMA1 in paleobotanical maize specimens found in Central Mexico could help determine if its genetic diversity was already reduced by 5,300 to 6,200 yr BP (Ramos-Madrigal et al., 2016 ; Vallebueno-Estrada et al., 2016; Benz, 2001). Our results open the possibility for environmental changes caused by volcanic activity acting as a driving force in the selection of loci involved in adaptive responses such as heavy metal tolerance.

Materials and Methods

Plant materials and growth conditions

Seeds from maize inbred line W22, teosinte parviglumis accession CIMMYTMA 29791, and insertional line LANMu1029790 were sterilized using a 1:1 distilled water and bleach solution, rinsed three times in water, and germinated in MS solid medium. One week-old seedings were transplanted into PVC tubes (length: 90 cm; diameter: 20 cm) or 20-liter pots filled with a pre-sterilized mix composed of 30% peat moss, 10% vermiculite, 10% perlite, and 50% ground litter, complemented with 15-9-12 (N-P-K) OsmocoteTM fertilizer. For HM growth conditions, 16 mg/kg of cadmium chloride (CdCl2) and 400 mg/kg of copper sulfate (CuSO4) were added to the substrate before seedling transplant. All plants were grown under glasshouse conditions at 28°C during day and 24°C during night, with relative humidity comprised between 40 and 50%.

Phenotypic analysis

All plants were harvested at 15 days (V2 stage), 30 days (V6 stage), and 50 days (R1 stage) after transplant. No less than ten plants were measured for each trait and stage. The width and length of immature leaves was measured in the 5th leaf. The length and width of mature leaves in the 3rd leaf. The total number of leaves, total number of stem internodes, plant height, basal stem diameter, total number of stem ramifications, and total fresh and dry weight of the aerial biomass were estimated at the R1 stage. Estimated root traits included the number and length of brace roots (aerial roots), the number and length of crown roots (underground roots originating within 2 cm below the basal node), number and length of nodal roots (underground roots originating more than 2 cm below the basal node), the number and length of seminal roots and maximum length of the root system.

Genetic diversity analysis

The SNP collection of maize chromosome 5 was analyzed on the basis of the HapMap V3.2.1 dataset (Bukowski et al. 2018), selecting individuals from maize landraces (LR) (Table S6) and teosinte parviglumis inbred line accessions (Teo; Table S7). A sliding window diversity (π) was calculated using VCFtools v0.1.13 (Danecek et al. 2011) with a window size of 100 bp (--window-pi 100) and a sliding step of 50 bp (--window-pi-step 50). Regions under selection were extracted from a list of 304 genes with domestication-related selection signatures across the whole chromosome (Hufford et al. 2012; Tables S8), for a total of 385,829 sliding windows (Table S9 and S10). A second set of neutral gene regions comprising 2,373,010 windows was also extracted across the entirety of chromosome 5. These two collections were compared to ZmHMA1 (GRMZM2G067853), ZmHMA7 (GRMZM2G029951), and ZmSKUs5 (GRMZM2G076985), encompassing 2,743 windows (Table S11 and S12). Additionally, we analyzed genetic diversity in the neighboring regions of these candidate genes, which included six genes (GRMZM2G701784, GRMZM2G014508, GRMZM2G367857, GRMZM2G375607, GRMZM2G374375, and GRMZM2G067883), totaling 3,088 windows. We employed a custom Perl script (Vallebueno GitHub repository) to calculate two diversity metrics: fold change (FC= π(LR)/π(Teo)) and the proportional diversity (FP=π(LR)/(π(LR) + π(Teo)). Data visualization was performed using violin plots generated in R v4.2.2.

For an arbitrary locus encompassing 30 Kbs up and downstream of the coding sequence of ZmMA1, ZmHMA7, and ZmSKUs5, the number of segregating sites (S), the number of unique sequences (haplotypes, h), and the nucleotide diversity index (θ) were estimated using DnaSP version 6.0 for all accessions included in HapMap3 (Rozas et al. 2017; Bukowski et al. 2018). DnaSP was also used to estimate Tajima’s D, Fu and Li’s D, and Fu and Li’s F indexes (Tajima 1989; Fu and Li 1993). The average proportion of pairwise nucleotide differences per nucleotide site (π) was calculated with VCFtools using bins of 100 bp and steps of 25 bp (Tajima 1983; Weir and Cockerham 1984). Null hypothesis of equality of parameters followed two-tailed tests as previously described (Rozas et al. 2017).

Gene expression analysis

Leaves, stems, tassels, female inflorescences, as well as primary, nodal, and crown roots were manually isolated, frozen in liquid nitrogen and stored at 4°C. Total RNA was extracted using TRIZOL (Thermo Fisher Scientific, USA). For reverse-transcriptase PCR (RT-PCR), RNA was converted to coding DNA using the One-step RT-PCR kit (QIAGEN, USA). The full list of primers is presented in Table S8. ZmCDK was used as a positive control. PCR amplification was conducted using the following parameters: at 94°C for 5 min.; 94°C 30 sec., 60°C 30 sec., 72°C 1 min. for 35 cycles; final stop after 72°C for 10 min. For quantitative real-time PCR (qPCR), total RNA was isolated from 4-week-old meristems using TRIZOL. Complementary DNA (cDNA) was synthesized using 4 μg of total RNA, 10 mM oligodTs and Superscript reverse transcriptase II (Invitrogen, USA). Primers for PCR were designed using the online program Primer 3 Plus (v.0.4.0) and verified with Oligocalculator (Sigma Chemical, St Louis MO) to discard dimer formation. PCR efficiencies of target and reference gene were determined by generating standard curves based on serial dilutions prepared for cDNA templates. PCR efficiency was calculated according to the slope of the standard curve (primers with 100% efficiency, the fold equals 2). Each quantitative real-time PCR (qPCR) reaction was performed in a 10 μl volume consisting of 5 μl of 23 SYBR Green PCR reaction Mix (Applied Biosystems, Foster City, CA). 3.5 μl of the DNA template (100 ng/ml), 0.5 μl of the forward primer (5 mM), 0.5 μl of the reverse primer (5 mM), and 0.5 μl of ultrapure water. The qPCR reactions were performed using the CFX96 Touch Real-Time PCR detection System and the data were analyzed using the Bio-Rad CFX Manager software v3.1. The thermal profile consisted of 10 min at 95°C, 40 cycles of 15 sec at 95°C, and 1 min at 60°C. Amplification results were collected at the end of the extension step. Primer sequences for qPCR amplification are listed in Table S7. A comparative 2-ΔΔCT method was used for determining a relative target quantity of gene expression (Schefe et al. 2006), and the cyclin-dependent kinase gene ZmCDK (GRMZM2G149286) was used as a control (Hu et al. 2019). Reproducibility of the results was evaluated for each sample by running three technical and three biological replicates of each of the reactions and each sample.

Statistical Analysis

Statistical analyses were conducted using RStudio, version 4.3.0 (Horton and Kleinman 2015). The data are presented as mean ± standard deviation and were tested by a paired Student’s T-Test. When the effects were significant according to T-Test, the treatments were compared with the Welch two sample T-Test at P < 0.05.

eLife Assessment

This study presents a valuable investigation into how heavy metal stress may have influenced the domestication of maize from its wild ancestor, teosinte parviglumis, focusing on specific ATPase genes with proposed roles in heavy metal homeostasis. The evidence supporting the main claims is incomplete, with suggestive but not definitive data linking gene function to domestication traits, and limited environmental context for the hypothesized selection pressures. While the work introduces an interesting model connecting environmental stress responses to evolutionary transitions and highlights underexplored aspects of teosinte plasticity, the conclusions would benefit from more comprehensive analyses such as transcriptomics, a broader survey of loci, and stronger paleoenvironmental validation. The study will be of interest to researchers in plant evolution and domestication, but currently lacks the analytical depth to fully support its central hypothesis.

Reviewer #1 (Public review):

In this study, Acosta-Bayona et al. aim to better understand how environmental conditions could have influenced specific gene functions that may have been selected for during the domestication of teosinte parviglumis into domesticated maize. The authors are particularly interested in identifying the initial phenotypic changes that led to the original divergence of these two subspecies. They selected heavy metal (HM) stress as the condition to investigate. While the justification for this choice remains speculative, paleoenvironmental data would add value; the authors hypothesize that volcanic activity near the region of origin could have played a role.

The authors exposed both maize and teosinte parviglumis to a fixed dose of copper and cadmium, representing an essential and a non-essential element, respectively. They assessed shoot and root phenotypic traits at a defined developmental stage in plants exposed to HM stress versus controls. They then focused on three genes already known to help plants manage HM stress: ZmHMA1, ZmHMA7, and ZmSKUs5. Two of these genes are located in a genomic region linked to traits selected during domestication. A closer examination of nucleotide variability in the coding and flanking regions of these genes provided evidence of selective pressure among teosinte parviglumis, maize, and the outgroup Tripsacum dactyloides.

They further generated a null mutant for ZmHMA1 and showed, for the first time in maize, a pleiotropic phenotype reminiscent of traits associated with the domestication syndrome. Finally, using qPCR, they reported increased expression of the domestication gene Teosinte branched1 (tb1) in teosinte parviglumis under HM stress. Comparative studies focusing on teosinte parviglumis and the genes ZmHMA1, ZmHMA7, and ZmSKUs5 under HM stress are limited; thus, this phenotypic characterization provides a promising starting point for further understanding the genetic basis of the response.

The dataset is of good quality, but the conclusions are not sufficiently supported by the data. Analyses should be expanded, and additional experiments included to strengthen the findings.

(1) Although the paper presents some interesting findings, it is difficult to distinguish which observations are novel versus already known in the literature regarding maize HM stress responses. The rationale behind focusing on specific loci is often lacking. For example, a statistically significant region identified via LOD score on chromosome 5 contains over 50 genes, yet the authors focus on three known HM-related genes without discussing others in the region. It is unclear why ZmHMA1 was selected for mutagenesis over ZmHMA7 or ZmSKUs5.

(2) The idea that HM stress impacted gene function and influenced human selection during domestication is of interest. However, the data presented do not convincingly link environmental factors with human-driven selection or the paleoenvironmental context of the transition. While lower nucleotide diversity values in maize could suggest selective pressure, it is not sufficient to infer human selection and could be due to other evolutionary processes. It is also unclear whether the statistical analysis was robust enough to rule out bias from a narrow locus selection. Furthermore, the addition of paleoclimate records (Paleoenvironmental Data Sources as a starting point) or conducting ecological niche modeling or crop growth models incorporating climate and soil scenarios would strengthen the arguments.

(3) Despite the interest in examining HM stress in maize and the presence of a pleiotropic phenotype, the assessment of the impact of gene expression is limited. The authors rely on qPCR for two ZmHMA genes and the locus tb1, known to be associated with maize architecture. A transcriptomic analysis would be necessary to 1- strengthen the proposed connection and 2- identify other genes with linked QTLs, such as those in the short arm of chromosome 5.

Reviewer #2 (Public review):

Summary:

This work explores the phenotypic developmental traits associated with Cu and Cd responses in teosinte parviglumis, a species evolutionary related to extant maize crops. Cu and Cd could serve as a proxy for heavy metals present in the soils. The manuscript explores potential genetic loci associated with heavy metal responses and domestication identified in previous studies. This includes heavy metal transporters, which are unregulated during stress. To study that, the authors compare the plant architecture of maize defective in ZmHMA1 and speculate on its association with domestication.

Strengths:

Very few studies covered the responses of teosintes to heavy metal stress. The physiological function of ZmHMA1 in maize also gives some novelty in this study. The idea and speculation section is interesting and well-implemented.

Weaknesses:

The authors explored Cu/Cd stress but not a more comprehensive panel of heavy metals, making the implications of this study quite narrow. Some techniques used, such as end-point RT-PCR and qPCR, are substandard for the field. The phenotypic changes explored are not clearly connected with the potential genetic mechanisms associated with them, with the exception of nodal roots. If teosintes in response to heavy metal have phenotypic similarity with modern landraces of maize, then heavy metal stress might have been a confounding factor in the selection of maize and not a potential driving factor. Similar to the positive selection of ZmHMA1 and its phenotypic traits. In that sense, there is no clear hypothesis of what the authors are looking for in this study, and it is hard to make conclusions based on the provided results to understand its importance. The authors do not provide any clear data on the potential influence of heavy metals in the field during the domestication of maize. The potential role of Tb-1 is not very clear either.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1(Public review):

In this study, Acosta-Bayona et al. aim to better understand how environmental conditions could have influenced specific gene functions that may have been selected for during the domestication of teosinte parviglumis into domesticated maize. The authors are particularly interested in identifying the initial phenotypic changes that led to the original divergence of these two subspecies. They selected heavy metal (HM) stress as the condition to investigate. While the justification for this choice remains speculative, paleoenvironmental data would add value; the authors hypothesize that volcanic activity near the region of origin could have played a role.

The justification of choice to investigate the effects of heavy metal stress is not speculative. As mentioned now in the Abstract, the elucidation of the genome from the Palomero toluqueño maize landrace revealed heavy metal effects during domestication (Vielle-Calzada et al., Science 2009). Our aim was to test the hypothesis that heavy metal (HM) stress influenced the evolutionary transition of teosinte parviglumis to maize.

(1) Although the paper presents some interesting findings, it is difficult to distinguish which observations are novel versus already known in the literature regarding maize HM stress responses. The rationale behind focusing on specific loci is often lacking. For example, a statistically significant region identified via LOD score on chromosome 5 contains over 50 genes, yet the authors focus on three known HM-related genes without discussing others in the region. It is unclear why ZmHMA1 was selected for mutagenesis over ZmHMA7 or ZmSKUs5.

We appreciated the depth and value of this comment.

Maize phenotypic responses to sublethal concentrations to heavy metals – copper (Cu) and cadmium (Cd) in particular - are well characterized and published, and in agreement with our results. In the first section of the Results (pgs 7 and 8), we added pertinent references to clearly show which observations are already known. By contrast, teosinte parviglumis responses are in all cases novel. To our knowledge this is the first study that analyzed in detail the phenotypic response of teosinte to sublethal concentrations of heavy metals, specifically Cu and Cd. We have now emphasized the novelty of these observations (pg 8).

To address the fact that we only focused on three known HM-related genes without discussing others in the statistically significant region identified via LOD score on chr.5, we have added a full section that reads as follows (pgs. 11 to 13 of the new version):

“Large-scale genomic and transcriptomic comparisons indicate that many HM response genes were positively selected across the maize genome.

To expand the results well beyond the analysis of the three genes previously described, we performed a detailed analysis of genetic diversity across the 11.47 Mb genomic region comprised between Z_mSKUs5_ and ZmHMA1. This additional analysis reveals general tendencies in the quantity and nature of loci that were affected by positive selection during the teosinte parviglumis to maize transition in a region identified via LOD score on chr.5. We compared nucleotide variability by using 100 bp bins covering loci composed of two 30 Kb segments up and downstream of coding sequences, respectively, and the coding sequence itself, for 173 genes present within the genomic region comprised between ZmSKUs5 and ZmHMA (Figure S1 and Supplementary File 6). Two types of statistical tests (ANOVA and Wilcoxon) were applied to nucleotide variability comparisons using the entirety of each locus. The Benjamini-Hochber procedure allowed an estimation of the false discovery rate (FDR<0.05) to avoid type I errors (false positives). Although some individual loci appear as differently classified depending on the statistical test applied (22 out of 173 loci), the general differences in nucleotide variability are consistently maintained within the subregions described below. We found that 166 out of 173 loci show signatures of positive selection and are roughly organized in five independent subregions of variable length. The first six loci are consecutively ordered in a 402 Kb subregion that includes ZmSKUs5. A second group of 13 consecutive loci expands over a 1.44 Mb subregion that contains NRAMP ALUMINUM TRANSPORTER1, also involved in HM response through uptake of divalent ions. A third group of 17 consecutive loci expands over 1.28 Mb; eleven contain genes encoding for uncharacterized proteins. The fourth group is composed of 57 consecutive loci expanding over 3.22 Mb and contains genes encoding for DEFECTIVE KERNEL55, AUXIN RESPONSE FACTOR16, and peroxydases involved in responses to oxydative stress. The fifth group contains 12 consecutive loci expanding over 713 Kb and contains ZmHMA1. An additional segment of approximately 1.17 Mb and containing 25 consecutive loci that were positively selected expands away from the ZmSKUs5-ZmHMA1 segment; it also contains several genes encoding for peroxydases. Although multiple loci include genes that could be involved in abiotic stress and oxidative responses, these results suggest that multiple factors other than HM stress could have played a role in the evolutionary mechanisms that affected the genetic diversity of chr.5 during the teosinte parviglumis to maize transition.

To further analyze the possibility that HM response could have played a role in maize emergence and subsequent domestication, we analyzed large scale transcriptomic data corresponding to independent experiments aiming at understanding the response of maize roots to HM stress. Six available transcriptomes were selected for in-depth analysis because they presented a fold change strictly higher than 1, and their results were supported by false discovery rates (FDR<0.05). These six transcriptomes (Table S5) included HM response datasets corresponding to growth conditions that not only incorporated Cu, but also lead (Pb) and chromium (Cr) that were not included in the substrate of our experiments. Transcriptional profiles were obtained from roots of plants at different stages: maize seedlings (Shen et al., 2012; Gao et al., 2015; Zhang et al., 2024a), three week old plantlets (Yang et al., 2023), and plants at V2 stage (Zhang et al., 2024b; Fengxia et al., 2025). A total of 120 genes shared by all six transcriptomes were found to be differentially expressed under HM stress conditions (66 upegulated and 54 downregulated; Figure S3), including ZmSKUs5, ZmHMA1 and ZmHMA7; 52 of them (43.3%) are located in maize loci showing less than 70% of the nucleotide variability found in teosinte parviglumis, suggesting that they were affected by positive selection (Yamasaki et al., 2005; Supplementary File 7). Of 18 mapping in chr.5, twelve are within the 82 cM that fractionates into multiple QTLs under selection during the parviglumis to maize transition. Interestingly, five additional loci containing HM response genes completely lack SNPs within their total length in both parviglumis and maize, and 19 additional loci lack SNPs in at least one 30 Kb segment or their coding region (Supplementary File 7), suggesting the frequent presence of ultraconserved genomic regions in many loci containing HM response genes. When this same analysis was conducted in a set of loci comprising 63 genes previously identified as differentially expressed in response to abiotic stress not directly related to HM responses (hypoxia; nutritional deficiency; soil alkalinity; drought; soil salinity), 18 loci (28.6%) showed less than 70% of the nucleotide variability found in teosinte parviglumis. Only one of them maps in chr.5 and none contained segments or coding regions lacking SNPs in parviglumis or maize. These results suggest that in contrast to other types of abiotic stress response genes, loci comprising a large set of genes that unambiguously respond to HM stress caused by chemical elements of diverse nature were affected by positive selection during the parviglumis to maize transition, irrespectively of their position in the genome.”

The detailed analysis of genetic diversity across 11.47 Mb of chr.5 in the genomic region comprised between ZmSKUs5 and ZmHMA1 in presented as Supplementary File 6.

The analysis of genetic diversity in loci encompassing heavy metal response genes shared by six transcriptomes and abiotic stress controls are described in Supplementary File 7.

In the Discussion (pgs. 21 and 22), we added a paragraph section that reads as follows:

“Although loss of genetic diversity is usually the result of human selection during domestication, it can also represent a consequence of natural selective pressures favoring fitness of specific teosinte parviglumis allelic variants better adapted to environmental changes and subsequently affected by human selection during the domestication process. This possibility is reflected by widely spread selective sweeps affecting a large portion of chr.5 that contains hundreds of genes showing signatures of positive selection. The analysis of 11.47 Mb covering the ZmHMA1ZmSKUs5 segment confirms the presence of large but discrete genomic subregions that were positively selected during the teosinte parviglumis to maize transition. Although several contain genes involved in HM response and oxidative stress, the diversity of gene functions does not necessarily favor abiotic stress over other factors that could be at the origin of selective forces affecting these regions. By contrast, a large scale transcriptomic survey indicates that genes consistently responding to HMs (Cu, Cd, Pb and Cr ) show signatures of positive selection at unusual high frequencies (43.3%) as compared to loci containing genes responding to other types of abiotic stress (28.6%). Our identification of HM response genes affected by positive selection is far from being exhaustive. Nevertheless, it agrees with the expected effects of a widespread selective sweep caused by environmental changes that influenced the parviglumis to maize transition at the genetic level. Of intriguing interest are 24 loci that partially or completely lack SNPs in both teosinte parviglumis and maize, suggesting possible genetic bottlenecks occurred before the teosinte to maize transition. Examples of other edaphological factors driving genetic divergence either in the teosintes or maize include local adaptation to phosphorus concentration in mexicana and parviglumis (Aguirre-Liguori et al. 2019), and fast maize adaptation to changing iron availability through the action of genes involved in its mobilization, uptake, and transport (Benke and Stich 2011). Our results reveal a teosinte parviglumis environmental plasticity that could be related to the function of HM response genes positively selected during the teosinte parviglumis to maize transition. Previous studies have demonstrated that transposable elements (TEs) contribute to activation of maize genes in response to abiotic stress, affecting up to 20% of the genes upregulated in response to abiotic stress, and as many as 33% of genes that are only expressed in response to stress (Makarevitch et al., 2015). It is therefore possible that the HM response of some specific genes that influenced maize emergence or domestication could be mediated by TEs influencing or driving their transcriptional regulation.”

The mutagenic analysis of ZmHMA7 and ZmSKUs5 will be included in a different publication.

(2) The idea that HM stress impacted gene function and influenced human selection during domestication is of interest. However, the data presented do not convincingly link environmental factors with human-driven selection or the paleoenvironmental context of the transition. While lower nucleotide diversity values in maize could suggest selective pressure, it is not sufficient to infer human selection and could be due to other evolutionary processes. It is also unclear whether the statistical analysis was robust enough to rule out bias from a narrow locus selection. Furthermore, the addition of paleoclimate records (Paleoenvironmental Data Sources as a starting point) or conducting ecological niche modeling or crop growth models incorporating climate and soil scenarios would strengthen the arguments.

We think that the detailed analysis of genetic diversity across 11.46 Mb covering the ZmSKUs5 to ZmHMA1 genomic segment – and its statistical validation - provides a precise understanding of the selective sweep dimensions in chr.5.

We do agree that lower nucleotide diversity values in maize are not sufficient to infer human selection. Because many HM response loci show unusually low nucleotide variability in teosinte parviglumis (see the results of the transcriptomic analysis presented above), we cannot discard the possibility that natural selection forces related to environmental changes could have affected native populations of teosinte parviglumis.

To further explore the link between environmental factors, natural or human-driven selection, and the paleoenvironmental context of the parviglumis to maize transition, we revised paleoenvironmental and geological records and added results in two sections that read as follows (pgs. 17 to 20):

“Paleoenvironmental studies reveal periods of climatic instability in the presumed region of maize emergence during the early Holocene.

It is well accepted that temperature fluctuations, volcanism and anthropogenic impact shaped the distribution and abundance of plant species in the Transmexican Volcanic Belt (TMVB) during the last 14,000 years (Torrescano-Valle et al. 2019). The TMVB has produced close to 8000 volcanic structures (Ferrari et al., 2011), transforming the relief multiple times, and causing hydrographic and soil changes that actively modified the distribution and composition of plant communities in Central Mexico. Detailed paleoenvironmental data for the Pleistocene and Holocene is available for several lacustrine zones located within the 50 to 100 km range of the region currently considered the cradle of maize domestication (Matzuoka et al. 2002; Figure 5a). In Lake Zirahuén (102°44′ W; 19°26′ N and approximately 2075 meters above sea level; index [i] in Figure 5a), pollen, microcharcoal and magnetic susceptibility analyses of two sedimentary sequences reveals three periods of major ecological change during the early and middle Holocene.

Between 9500 and 9000 calibrated years before present (cal yr BP), pine forests seem to have been associated with summer insolation increases. A second peak of forest change occurred at around 8200 cal yr BP, coinciding with cold oscillations documented in the North Atlantic. Finally, events occurred between 7500 and 7100 cal yr BP shows an abrupt change in the plant community related to humid Holocene climates and a presumed volcanic event (Lozano-García et al., 2013). The environmental history of the central Balsas watershed has also been documented by pollen, charcoal, and sedimentary analysis conducted in three lakes and a swamp of the Iguala valley (Piperno et al. 2007). Paleoecological records of lake Ixtacyola (8°20N, 99°35W and approximately 720 meters above sea level; index [ii] in Figure 5a) and lake Ixtapa (8°21N, 99°26W) indicate that an important increase in temperature and precipitation occurred between 13000 and 10000 cal yr BP. The pollen record of Ixtacyola showed that members of the genus Zea were already part of the vegetation coverage by 12900 to 13000 cal yr BP, suggesting that some teosintes – likely including parviglumis - were commonly found at elevation areas where they do not presently occur. Lake Almoloya (also named Chignahuapan; 19°05N, 99°20E and approximately 2575 meters above sea level; index [iii] in Figure 5a) in the upper Lerma basin is only 20 Km from the crater of the Nevado de Toluca that is responsible for creating the late Pleistocene Upper Toluca Pumice layer over which the Lerma basin is deposited. Pollen records indicate the presence of Zea species by 11080 to 10780 cal yr BP. As for other locations, an important period of climatic instability prevailed between 11500 and 8500 cal yr BP (Ludlow-Wiechers et al., 2005). Humidity fluctuations occurred until 8000 cal yr BP, with a stable temperate climate between 8500 and 5000 cal yr BP. Although pollen and diatom studies are often difficult to interpret at a regional scale, the overall results presented above suggest consistent periods of Zea plants present in periods of environmental and climatic instability that correlate with the history of volcanic activity during the early Holocene, as described in the next section.

Temporal and geographical convergence between volcanic eruptions and maize emergence during the Holocene.

Current evidence indicates that the emergence and domestication of maize initiated in Mesoamerica some time around 9,000 yr BP (Matsuoka et al. 2002). The current location of teosinte parviglumis populations that are phylogenetically most closely allied with maize are currently distributed in a region located between the Michoacan-Guanajuato Volcanic Field (MGVF) at their northwest, and the Nevado de Toluca and Popocatéptl volcanoes at their east and northeast (Figure 5a; Matsuoka et al. 2002). Precise records of field data indicate that ten accessions were collected in the Balsas river drainage near Teloloapan and Sierra de Huautla (Guerrero), at approximately 100 km south of the Nevado de Toluca crater. Three other accessions were collected near Tejupilco de Hidalgo and Zacazonapan (Estado de México), at approximately 50 to 60 km from the Nevado de Toluca crater (8762, JSG y LOS-161, and JSG-391). And four other accessions were located in Michoacan, at a location within the MGVF (accession 8763), or at mid-distance between the MGVF and the Nevado de Toluca crater (accessions JSG y LOS-130, 8761, and 8766).

The most important source of HMs in ancient soils of Mesoamerica is TMBV-dependent volcanic activity through short- and long-term effects related to lava deposits, ores, hydrothermal flow, and ash (Torrescano-Valle et al. 2019). The Nevado de Toluca volcano produced one of the most powerful eruptions from central Mesoamerica in the Holocene, giving rise to the Upper Toluca Pumice deposit at 12621 to 12025 cal yr BP (Arce et al., 2003; Figure 5b). The pumice fallout blanketed the Lerma and Mexico basins with 40 cm of coarse ash (Bloomfield and Valastro 1977; Arce et al. 2003). A second eruption dated by 36Cl exposure occurred at 9700 cal yr BP (Arce et al. 2003; Figure 5b), and the most recent eruption occurred at 3580 to 3831 cal yr BP (Macías et al. 1997). During the early and middle Holocene, the Popocatéptl volcano produced at least four eruptions dated 13037-12060, 10775–9564, 8328-7591, and 6262-5318 cal yr BP (Siebe et al. 1997); three other important eruptions occurred during the late Holocene, between 2713 and 733 cal yr BP (Siebe and Macías, 2006). In addition, the MGFV is a monogenetic volcanic field for which 23 independent eruptions have been documented during the Holocene, 21 of them located towards the southern part of the field, in close proximity to the region harboring some of the teosinte parviglumis populations most closely related to maize. Three of these eruptions occurred in the early Holocene (El Huanillo 1130 to 9688 cal yr BP; La Taza 10649 to 10300 cal yr BP; Cerro Grande 10173 to 9502 cal yr BP; Figure 5b), and three others during the initial period of the middle Holocene, between 8400 and 7696 cal yr BP (La Mina, Los Caballos, and Cerro Amarillo; Figure 5b). On average, a new volcano forms every ~435 years in the MGFV (Macías and Arce, 2019). No less than 16 other eruptions occurred between 7159 cal yr BP and the present time (Figure 5b). Soils of volcanic origin (andosols) are currently distributed in regions north-west from the Nevado de Toluca and Popocatéptl craters, in close proximity with teosinte parviglumis populations most closely related to maize (Figure S5). Although modern distribution of teosinte populations may differ from their distribution around 9000 yr BP, and unknown populations more closely related to maize may yet to be discovered, this data indicates that the date and region where maize emerged is convergent with the dates and locations of several volcanic eruptions occurred during the Holocene in that same region.”

(3) Despite the interest in examining HM stress in maize and the presence of a pleiotropic phenotype, the assessment of the impact of gene expression is limited. The authors rely on qPCR for two ZmHMA genes and the locus tb1, known to be associated with maize architecture. A transcriptomic analysis would be necessary to 1- strengthen the proposed connection and 2- identify other genes with linked QTLs, such as those in the short arm of chromosome 5.

Real-time qPCR is an accurate and reliable approach to assess the expression of specific genes such as ZMHMA1 and Tb1, but we agree that our results do not allow to establish a direct regulatory link between the function of Tb1, the pleiotropic parviglumis phenotype under HM stress, and the function of ZmHMA1. We also concede that the large transcriptional analysis of HM response in maize (presented above) does not allow to elucidate a possible connection between these two genes. We have substantially downplayed our conclusion in this section by modifying the end of the section in pg. 17, that now reads:

“These results do not allow to directly link the regulation of ZmHMA1 expression to the function of Tb1; however, they open an opportunity to further investigate the possibility that under HM stress, the formation of secondary ramifications in teosinte parviglumis could be repressed by transcription factors of the TCP family, including Tb1.”

This is also emphasized in the Discussion (pg 21) as follows:

“Under HM stress, we also show that Tb1 is overexpressed in the apical meristem of teosinte parviglumis, suggesting that formation of secondary ramifications is repressed by Tb1 function under HM stress, as in extant maize. At this stage we cannot discard the possibility that Tb1 upregulation in parviglumis reflects a more generalized response to abiotic stress; however, the expression ZmHMA1 is downregulated in W22 wild-type maize meristems in the presence of HMs but upregulated in teosinte parviglumis meristems, suggesting that a specific regulatory shift relating HM responses and ZmHMA1 function occurred during the teosinte parviglumis to maize transition.”

On the other hand, the transcriptional analysis the identification of 52 additional HM response genes showing signatures of positive selection occurred during the parviglumis to maize transition; 12 of them map to chr.5 within the region having linked QTLs within the short arm of chr.5. So far, genes involved in HM response and oxidative stress represent the most prevalent class of genes identified within the genomic region showing pleiotropic effects on domestication and multiple linked QTLs in chr.5.

Reviewer #2 (Public review):

Summary:

This work explores the phenotypic developmental traits associated with Cu and Cd responses in teosinte parviglumis, a species evolutionary related to extant maize crops. Cu and Cd could serve as a proxy for heavy metals present in the soils. The manuscript explores potential genetic loci associated with heavy metal responses and domestication identified in previous studies. This includes heavy metal transporters, which are unregulated during stress. To study that, the authors compare the plant architecture of maize defective in ZmHMA1 and speculate on its association with domestication.

Strengths:

Very few studies covered the responses of teosintes to heavy metal stress. The physiological function of ZmHMA1 in maize also gives some novelty in this study. The idea and speculation section is interesting and well-implemented.

Weaknesses:

The authors explored Cu/Cd stress but not a more comprehensive panel of heavy metals, making the implications of this study quite narrow. Some techniques used, such as end-point RT-PCR and qPCR, are substandard for the field. The phenotypic changes explored are not clearly connected with the potential genetic mechanisms associated with them, with the exception of nodal roots. If teosintes in response to heavy metal have phenotypic similarity with modern landraces of maize, then heavy metal stress might have been a confounding factor in the selection of maize and not a potential driving factor. Similar to the positive selection of ZmHMA1 and its phenotypic traits. In that sense, there is no clear hypothesis of what the authors are looking for in this study, and it is hard to make conclusions based on the provided results to understand its importance. The authors do not provide any clear data on the potential influence of heavy metals in the field during the domestication of maize. The potential role of Tb-1 is not very clear either.

Thank you for these comments. We have now emphasized our hypothesis in the abstract and the last paragraph of the Introduction (pg. 6):

“To test the hypothesis that heavy metal (HM) stress influenced the evolutionary transition of teosinte to maize, we exposed both subspecies to sublethal concentrations of copper and cadmium etc…”

A comprehensive panel of heavy metals would not be more accurate in terms of simulating the composition of soils evolving across 9,000 years in the region where maize presumably emerged. Copper (Cu) and cadmium (Cu) correspond each to a different affinity group for proteins of the ZmHMA family. ZmHMA1 has preferential affinity for Cu and Ag (silver), whereas ZmHMA7 has preferential affinity to Cd, Zn (zinc), Co (cobalt), and Pb (lead). Since these P1b-ATPase transporters mediate the movement of divalent cations, their function remains consistent regardless of the specific metal tested, provided it belongs to the respective affinity group. By applying sublethal concentrations of Cd (16 mg/kg) and Cu (400 mg/kg), we caused a measurable physiological response while allowing plants to complete their life cycle, including the reproductive phase, facilitating a comprehensive analysis of metal stress adaptation. Whereas higher doses impair flowering or are lethal, lower Cu/Cd concentrations do not consistently show conventional phenotypic responses such as reduced plant growth (AbdElgawad et al. 2020; Atta et al., 2023)

Based on comments by both reviewers, we present now a large transcriptional analysis that incorporates HM responses to lead (Pb) and chromium (Cr), in addition to Cu. Results show that many genes responding to Pb and Cr were also positively selected across the maize genome, suggesting that HM stress led to a ubiquitous rather than a specific evolutionary response to heavy metals (please see our response to Reviewer#1 and sections in pgs. 11 to 13) .

Real-time qPCR is an accurate and reliable approach to assess the expression of specific genes such as ZMHMA1 and Tb1, but we agree that our results do not allow to establish a direct regulatory link between the function of Tb1, the pleiotropic parviglumis phenotype under HM stress, and the function of ZmHMA1. We also concede that the large transcriptional analysis of HM response in maize (presented above) does not allow to elucidate a possible connection between these two genes. Therefore, we have substantially downplayed our conclusion in this section by modifying the end of the section in pg. 17, that now reads:

“These results do not allow to directly link the regulation of ZmHMA1 expression to the function of Tb1; however, they open an opportunity to further investigate the possibility that under HM stress, the formation of secondary ramifications in teosinte parviglumis could be repressed by transcription factors of the TCP family, including Tb1.”

There are two phenotypic changes clearly connected with the genetic mechanisms involved in the parviglumis to maize transition: plant height and the number of seminal roots (not nodal roots). These changes have been now emphasized in the Abstract and the description of the results.

Regarding the possibility for HM stress to represent a confounding factor in the selection of maize and not a driving factor, we expanded the genomic analysis of genetic diversity well beyond the analysis of the three genes under initial study, to cover a segment of 11.47 Mb comprised between ZmSKUs5 and ZmHMA1. We compared nucleotide variability by using 100 bp bins covering loci composed of two 30 Kb segments up and downstream of coding sequences, respectively, and the coding sequence itself, for 173 genes present within the genomic region comprised between ZmSKUs5 and ZmHMA (Figure S1 and Supplementary File 6). The full analysis is presented in a new section pgs. 11 and 12. We found that 166 out of 173 loci show signatures of positive selection and are roughly organized in five independent subregions of variable length. Four out of five subregions contain more than one HM or oxidative stress response gene within loci showing signatures of positive selection. Although multiple factors other than HM stress could have played a role in the evolutionary mechanisms that affected the genetic diversity of chr.5, large scale transcriptomic data corresponding to independent experiments aiming at understanding the response of maize roots to HM stress allowed the identification of 49 additional HM response genes within loci showing positive selection across the genome, a proportion (43.3%) far greater than the proportion of loci containing response genes to other types of abiotic stress not related to HMs (28.6%). These results are described in detail in pgs. 12 and 13 (Figure S3 and Supplementary File 7). These results provide strong evidence in favor of HM stress and not another factor driving positive selection.

We now provide precise and pertinent paleoenvironmental data on the potential influence of heavy metals in the field. In sections pgs. 17 to 20 we review paleoenvironmental studies revealing periods of climatic instability in the presumed region of maize emergence during the early Holocene, and data indicating that the date and region where maize emerged is convergent with the dates and locations of several volcanic eruptions occurred during the early and middle Holocene in that same region. Please see responses to Reviewer#1 for details.

We agree that our results do not allow to establish a direct regulatory link between the function of Tb1, the pleiotropic parviglumis phenotype under HM stress, and the function of ZmHMA1. We also concede that the large transcriptional analysis of HM response in maize (presented above) does not allow to elucidate a possible connection between these two genes. Therefore, we have substantially downplayed our conclusion in this section by modifying the end of the section in pg. 17, that now reads:

“These results do not allow to directly link the regulation of ZmHMA1 expression to the function of Tb1; however, they open an opportunity to further investigate the possibility that under HM stress, the formation of secondary ramifications in teosinte parviglumis could be repressed by transcription factors of the TCP family, including Tb1.”

This is also emphasized in the Discussion (pg 21) as follows:

“Under HM stress, we also show that Tb1 is overexpressed in the apical meristem of teosinte parviglumis, suggesting that formation of secondary ramifications is repressed by Tb1 function under HM stress, as in extant maize. At this stage we cannot discard the possibility that Tb1 upregulation in parviglumis reflects a more generalized response to abiotic stress; however, the expression ZmHMA1 is downregulated in W22 wild-type maize meristems in the presence of HMs but upregulated in teosinte parviglumis meristems, suggesting that a specific regulatory shift relating HM responses and ZmHMA1 function occurred during the teosinte parviglumis to maize transition.”

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

While the dataset generated provides an interesting foundation for hypothesis testing on HM stress and domestication, the current data do not sufficiently support the conclusions of the manuscript.

(1) The description of maize and teosinte architecture under HM stress is well presented.

However, traits like shoot height, leaf size reduction, and biomass loss also occur under other environmental stresses such as drought and salinity. Additional evidence beyond shoot and root architecture would help validate the link between tb1 expression and specific ZmHMA genes under HM stress, or whether it reflects a more generalized stress response.

We have already addressed in detail this point in the public response to Reviewer#1.

(2) The nucleotide variability analysis is interesting, but I would have liked to see additional information to clarify the choice of the data selection and the strength of the conclusions with human selection.

We have already addressed in detail this point in the public response to Reviewer#1.

a) The choice of Tripsacum dactyloides as the outgroup to determine nucleotide variability seems to be distant, and I wonder whether other combinations with a closer outgroup or multiple outgroups were tried to provide a more accurate context.

Nucleotide variability in Tripsacum dactyloides is used to graphically illustrate an external reference and not as an outgroup in the extended analysis of genetic diversity at the locus and genomic level. We did not used Tripsacum dactyloides as an outgroup in our statisticalm analysis. We could have indeed a closer teosinte subspecies as an outgroup, but at this stage no data warrants that environmentally-related selective pressures could have affected genetic diversite in other teosintes. This possibility in currently being investigated.

b) Evolutionary differences not related to human influence could affect the results. The phrase "order of magnitude difference in π values" needs statistical validation (e.g., confidence intervals, p-values).

We agree and have eliminated the sentence, as it is no longer relevant at the light of the detailed genomic analysis of genetic diversity prsented in Supplementary File 6.

c) The comparison with ZmGLB1, a neutral control locus, suggests that domestication-related changes in nucleotide variability are specific to the three candidate genes. However, the concept of neutrality is complex, and while ZmGLB1 may be considered neutral in this case, the argument does not address the possibility of other factors, such as linked selection, that could influence variability in these genes. Referencing Hufford et al. is insufficient and would require a deeper argument.

We also agree with this comment. We think that the influence and consequences of linked selection are now well documented for 11.46 Mb analyzed in chr.5 (pgs 11 and 12) in the main text and Supplementary File 6).

(3) The statement: "Our evidence indicates that HM stress revealed a teosinte parviglumis environmental plasticity that is directly related to the function of specific HM response genes that were affected by domestication through human selection" is not supported by the presented data. The rationale for the specific Cd/Cu dosage used is unclear. A dose-response gradient would better demonstrate the nature and strength of the plastic response.

Previous reports support the rationale for the specific HM dosage in this study; Cu/Cd dosage response gradients have been conducted in maize (AbdElgawad et al. 2020; Atta et al., 202), but since no studies have been conducted in teosinte, we reasoned that it was important to apply the same treatment to both subspecies. We have now emphasized this rationale by adding the following in pg XX: “Whereas higher doses impair flowering or are lethal, lower Cu/Cd concentrations do not consistently show conventional phenotypic responses such as reduced plant growth (AbdElgawad et al. 2020; Atta et al., 2023)”.

We agree that the statement raised by the reviewer needed revision at the light of our results. We did revise the statement to accurately reflect our current evidence as follows: “Our results reveal a teosinte parviglumis environmental plasticity that is likely related to the function of HM response genes positively selected during the teosinte parviglumis to maize transition.”

(4) In maize, TEs are known to influence gene expression under abiotic stress, including for tb1 (PMID: 25569788). Since the author appears to make a causative conclusion between ZmHMA1, TB1, and HM stress, I would have liked to see a whole-transcriptome analysis and not a curation of two genes to determine whether other factors, such as TEs, can have that would lead to similar outcomes.

We agree that is definetely a possibility that we have not investigated at this stage. However, we added a pargraph to reflect this pertinent suggestion:

“Previous studies have demonstrated that transposable elements (TEs) contribute to activation of maize genes in response to abiotic stress, affecting up to 20% of the genes upregulated in response to abiotic stress, and as many as 33% of genes that are only expressed in response to stress (Makarevitch et al., 2015). It is therefore possible that the HM response of some specific genes that influenced maize emergence or domestication could be mediated by TEs influencing or driving their transcriptional regulation.”

(5) I would suggest that the authors carefully review the tables, figures, and the corresponding legends. For example :

a) Table 2 is called before Table 1, I would therefore suggest changing the numbering to reflect the paragraph order.

Thank you for your help, we did change the order of the Tables in the new version.

b) In Table 2, it is not clear whether the P value applies to the mean difference between WT and the mutant zmhma1, either in the presence or the absence of heavy metals. In addition, the authors need to use the P-value to estimate the differences between WT in the absence vs presence of HM, and WT in the absence of HM versus the mutant in the absence of HM (idem for presence).

We did address this issue in detail and added P-values and specific pairwise comparisons to that Table (now Table 1). Data are presented as mean ± standard deviation and were tested by a paired Student’s T-Test. When the effects were significant according to T-Test, the treatments were compared with the Welch two sample T-Test at P < 0.05.

c) Table 1 and Table 2: Indicate what type of statistical test was used and the number of plants used for each experiment (n). Also, I recommend the use of scientific notation for the P-values.

The statistical tests have now been indicated, scientific notation has been added to the P-values; the number of plants and biological replicates are indicated in the Methods section.

d) Lines 202 and 204: I assume Table 1 should be called instead of Table 2.

This error has been corrected.

e) General: In the text, when significance is highlighted along with measurements, the p-value needs to be added.

We have added the P-value along the measurement for all significant differences.

f) In the text, it is also mentioned that "the expression of ZMHMA1 was significantly increased in the presence of HMs (Figure 3c)". We are looking here at an RT-PCR, which is qualitative and without a robust quantitative comparison and statistics, I cannot conclude this assessment based on the presented evidence. No statistical measure is indicated here.

Panel 3c is not RT-PCR but a real-time qPCR, showing relative fold-change, normalized to actin, with a 3-technical triplicate per 3 biological replicates). We have added error bars (SD) and P-values represented by asterisks (calculated with Student's t statistic) to support significant differences (P<0.05 and P<0.01). ZmHMA1 expression was significantly increased in the presence of HMs only in teosinte; there was no significant difference in maize.

g) Figure 3 should at least have the gene name in the figure to quickly understand the figure panel. The key conserved domains should also be identified.

We agree and apologize for the omission. The gene names have been added adjacent to the structures.

h) Sentence at lines 459-460 lacks words and punctuation.

This unfortunate rror has also been corrected.

i) Figure S1, the reference Lemmon and Doebley, 2024 should be Lemmon and Doebley, 2014 to harmonize with the text.

The correct year is 2014. We have corrected this error.

Reviewer #2 (Recommendations for the authors):

(1) The narrative should be clearer, starting with a clearer hypothesis that is later sustained or not in the results, and then discussed in the idea and speculation section.

Thank you for the comment. We have clarified the hypothesis, it is included in the abstract and the last paragraph of the Introduction. We hope it is now clear that the evidence presented supports our hypothesis

(2) Focus more on traits that are relevant, for example, nodal and seminal roots.

We modified the text to emphasize three relevant traits. In the case of teosinte under HM stress, absence of tillering and increase in the number of female inflorescences. In the case of the zmha1 mutant under HM stress, differences in the number of nodal roots, and differences in height.

(3) RNA-seq in Cu/Cd stress could make the work much more useful and complete.

As previously mentioned, we have incorporated a large scale transcriptional analysis on the basis of six transcriptomes statistically validated (Table S5). Please see sections pgs. 11 to 13 for details.

Version 2

Abstract

Maize originated from teosinte parviglumis following a subspeciation event occurred in volcanic regions of Mesoamerica. The elucidation of the phenotypic changes that gave rise to maize have focused on the direct consequences of domestication, with no insights on how environmental factors could have influenced specific gene function and human selection. The genome of the Palomero toluqueño landrace suggested heavy metal (HM) effects on maize domestication (Vielle-Calzada et al., 2009). To test the hypothesis that HM stress influenced the evolutionary transition of teosinte to maize, we exposed both subspecies to sublethal concentrations of copper and cadmium. We also assessed the genetic diversity of three HM response genes mapping to chromosome 5 (chr.5) and previously shown to be affected by domestication: ZmHMA1, ZmHMA7 – encoding for heavy metal ATPases of the P1b family-, and ZmSKUs5, encoding for a multicopper oxidase. ZmHMA1 and ZmSKUs5 are within a genomic region containing hundreds of genes linked to QTLs with pleiotropic effects on domestication. The genomic analysis of chr.5 shows that the three genes were under strong positive selection as compared to previously identified domestication genes. Large-scale transcriptomic comparisons indicate that many other loci containing HM response genes were also positively selected across the maize genome. Exposure of teosinte parviglumis to HM stress results in a plant architecture reminiscent of extant maize, and upregulation of Teosinte branched1 (Tb1) in the meristem. ZmHMA1 and ZmHMA7 are expressed throughout development and respond to HM stress in both subspecies. ZmHMA1 is mainly involved in restricting plant height and optimizing the number of female inflorescences and seminal roots. Paleoenvironmental studies reveal a temporal and geographical convergence between volcanic eruptions and maize emergence during the early Holocene, at periods of climatic instability. Overall, these results suggest that abiotic stress influenced the evolutionary transition that gave rise to extant maize through the activity of heavy metal response genes and their phenotypic effects, as a possible consequence of volcanic activity.

Introduction

Driven by either conscious or unconscious human selection, domestication can be considered a gradual evolutionary process resulting in the adaptation to agroecological environments and anthropogenic preferences (Ross-Ibarra et al. 2007). In the case of plants, the obvious consequences of domestication are represented by visually recognizable traits that include changes in photoperiod response, reduced capacity of seed dispersion, reduced lateral branching, larger fruits or grains, reduced duration of seed dormancy, and increase in the size of female inflorescences (Pickersgill 2007; Larson et al. 2014). Notably, research efforts to understand the phenotypic changes that gave rise to maize have mainly focused on the direct consequences of domestication in aerial organs and sometimes roots (Chen et al. 2022; Lopez-Valdivia et al. 2022; Ren et al. 2022), with no emphasis on the role of adaptive environmental responses that, by acting on specific gene function, could have given rise to phenotypic traits that were identified and selected by humans.

It is generally accepted that domestication of maize (Zea mays ssp. mays) from teosinte parviglumis (Zea mays ssp. parviglumis) initiated in Mesoamerica around 9,000 years before present (Matsuoka et al. 2002). The initial evolutionary transition that resulted in the event of subspeciation most likely occurred close to the Balsas river drainage, in regions geographically converging with the Trans-Mexican Volcanic Belt (TMVB), at the intersection of the States of Mexico, Guerrero, and Michoacan (Matsuoka et al. 2002). Although the nature of the initial phenotypic changes that allowed the emergence of primitive maize remains unclear, the most obvious architectural and morphological differences between both subspecies affect their vegetative and reproductive organs. Teosinte parviglumis generates numerous lateral branches ending in male inflorescences, while maize often has a few short or no lateral branches, not ending in male reproductive organs (Wills et al. 2013). Additionally, the seeds of teosinte parviglumis are covered by a hard glume that protects the kernel externally, while in maize glumes are reduced, soft and do not cover the kernel (Doebley and Stec 1993). Large scale efforts involving physical mapping and comparative genomic analysis have resulted in the identification of numerous regions showing differential nucleotide variability between teosinte parviglumis and the B73 reference maize genome. In some cases, mutations in genes located within those genomic regions result in phenotypes that reveal their involvement in domestication traits. One of the major genes affected by maize domestication is Teosinte branched1 (Tb1), a transcription factor of the TCP family that controls lateral branching, sexual determination, and glume formation (Doebley et al. 1997; Martín-Trillo and Cubas 2010). Tb1 acts as a repressor of lateral branching (Studer et al. 2017), giving rise to maize plants usually composed of a single stem. Other genes involved in domestication include Teosinte glume architecture1 (Tga1), ramosa2 (ra2) and Zea floricaula leafy2 (zfl2), among others (Bomblies and Doebley 2006; Sigmon and Vollbrecht 2010; Wang et al. 2015).

The importance of phenotypic plasticity is considered as likely relevant in the domestication history of crops (Ross-Ibarra et al. 2007; Lorant et al. 2020; Mueller et al., 2023); however, the role of environmental factors influencing their origins remains elusive and poorly investigated. A previous study showed that teosinte parviglumis grown under low temperatures and CO2 concentrations - reminiscent of those prevailing in the late Pleistocene and early Holocene - exhibit some phenotypes related to the domestication syndrome, including short branches, a single main stalk, and partially open seed fruitcases (Piperno et al. 2015; Piperno et al. 2019); however, the genetic basis of these changes have not been investigated. Although these results suggest that phenotypic plasticity could have influenced the domestication process, to this date there are no hints on how the adaptation to local environmental conditions could have affected specific gene function and guided human selection.

Recent genomic evidence indicates that admixture of maize with native teosinte mexicana populations (Zea mays ssp. mexicana) contributed to maize adaptation across a wide range of altitudes and latitudes in the American continent (Yang et al. 2023); however, the earliest maize macro-specimens found to date lack teosinte mexicana admixture and exhibit morphological attributes of fully domesticated cobs, indicating that mexicana is not a direct ancestor of maize and that its introgression occurred after initial divergence from teosinte parviglumis (Vallebueno-Estrada et al. 2023; Acosta-Bayona et al. 2024). Consequently, the nature of the genetic factors that guided the emergence of maize remain largely unknown, and the initial phenotypic changes that caused the original divergence of these two subspecies have not been identified. It is also unknown if domestication initiated by direct action of human selection on teosinte natural standing variation, or if the response to specific environmental factors generated phenotypic changes that were identified and selected by humans to act on native teosinte parviglumis populations.

Maize can adapt to abiotic factors such as heavy metal (HM) stress through numerous mechanisms that prevail during development (AbdElgawad et al. 2020; Gallo-Franco et al. 2020; Malenica et al. 2021; Gao et al. 2022). These include metal immobilization by mycorrhizal association, metal sequestration, or complexation by exuding organic compounds from roots, metal intracellular sequestration and compartmentalization in vacuoles. HMs include essential elements that are necessary for maize growth through enzymatic function, such as iron (Fe), manganese (Mn), zinc (Zn), magnesium (Mg), molybdenum (Mo), and copper (Cu); and non-essential elements like Cadmium (Cd), chromium (Cr), lead (Pb), aluminum (Al), and selenium (Se). In general, HM stress leads to visibly reduced plant growth due to the reduced cell elongation and cell wall elasticity. At the molecular level, metal toxicity can disturb the cellular redox balance and lead directly or indirectly to oxidative stress and the formation of reactive oxygen species (ROS), shifting the redox balance to the pro-oxidative side until the antioxidative defense system allows a new balanced redox status (Smeets et al. 2009; Hall 2002).

The comparison of the Palomero toluqueño and B73 genomes resulted in the identification of a large collection of identical sequence regions (IdSRs) with low nucleotide variability and transcriptional evidence of expression (Vielle-Calzada et al. 2009). Putative domestication loci included ZmHMA1 and ZmHMA7, two genes encoding members of the P1b-ATPases protein superfamily, also known as Heavy Metal ATPases (P-type ATPases) characterized by the generation of a phosphorylated intermediate during the transport reaction cycle. Members of this superfamily contain 8 to 12 transmembrane domains as well as conserved ATP binding, phosphorylation and dephosphorylation sites (Solioz and Vulpe 1996; Axelsen and Palmgren 2001; Voskoboinik et al. 2003; Williams and Mills 2005). The superfamily can be divided in subfamilies that include H±ATPases (type 3A) in plants and fungi, Na+/K ATPases (type 2C/D) in animals, Ca2±ATPases (type 2A/B), and heavy metal (HM) transporting ATPases (type 1B) capable of transporting a large variety of HM cations across cell membranes. ZmHMA1 and ZmHMA7 are type 1B ATPases that include a conserved Cys-Pro-Cys/His/Ser motif within a transmembrane domain necessary for the translocation of HMs cations through the membrane channel. A third Palomero toluqueño/B73 IdSR included Skewed Root Growth Similar5 (ZmSKUs5), a gene encoding a Skewed5 (SKU5)-similar (SKS) protein classified as a multi-copper oxidase containing cuperdoxin domains (Sedbrook et al. 2002). All three genes map to chromosome five (chr.5), with ZmHMA1 and ZmSKUs5 within a genomic region that fractionates into multiple QTLs under selection during the parviglumis to maize transition (Doebley and Stec 1993; Lemmon and Doebley, 2014). Contrary to major domestication loci showing a single highly pleiotropic gene responsible for important domestication traits, in this chr.5 genomic region the smallest 1.5-LOD support interval contained more than 50 genes, suggesting that phenotypic effects are due to multiple linked QTLs (Lemmon and Doebley, 2014). While the functional characterization of all three maize genes remains elusive, the Arabidopsis and barley ZmHMA1 homologs act during heavy metal transport in the chloroplast (Boutigny et al. 2014), demonstrating their functional involvement in homeostasis.

To investigate the possibility that HM response had an effect on the origin and domestication of maize, we compared the phenotypic changes that teosinte parviglumis and maize undergo when grown in the presence of sublethal concentrations of Cu and Cd. We show for the first time that teosinte parviglumis responds to HM stress by acquiring an aerial architecture reminiscent of extant maize and unusual proliferation of female inflorescences. Tb1 expression is upregulated in teosinte parviglumis meristems when grown under HM stress. Analysis of nucleotide variability across 11.46 Mb of chr.5 indicates that many genes within the ZmHMA1 to ZmSKUs5 genomic region show reduced nucleotide variability in maize as compared to teosinte parviglumis. Although the reduction is not specific to these genes, large-scale transcriptional analyses suggests that contrary to genes involved in other types of abiotic stress, a large number of loci that contain HM response genes were affected by positive selection during the teosinte parviglumis to maize transition. Both ZmHMA1 and ZmHMA7 are active throughout maize and teosinte parviglumis development, and their expression responds to HM stress. In addition, mutations in ZmHMA1 show domestication-related defects such as the absence of seminal roots and increased height, both manifested under exposure to HMs. Paleoenvironmental studies reveal a temporal and geographical convergence that relates the history of volcanic eruptions in the TMVB with the emergence of maize at times of climatic instability. Overall, this evidence indicates that the evolutionary transition that gave rise to extant maize was influenced by heavy metal stress through the activity of P1b-ATPases, presumably responding to edaphological consequences of volcanic activity.

Results

Teosinte parviglumis responds to heavy metal stress by acquiring an architecture reminiscent of extant maize

To compare HM response in teosinte parviglumis and maize, we exposed wild-type W22 maize inbred and CIMMYTMA 29791 teosinte parviglumis individuals to sublethal concentrations of Cu (400 mg/kg of CuSO4) and Cd (16mg/kg of CdCl2). CuSO4 and CdCl2 were homogenized in soil before transplant and grown under controlled greenhouse conditions. HM concentrations were based on preliminary dose responses that allowed maize plants to reach reproductive stages and complete their life cycle (AbdElgawad et al. 2020). Whereas higher doses impair flowering or are lethal, lower Cu/Cd concentrations do not consistently show conventional phenotypic responses such as reduced plant growth (AbdElgawad et al. 2020; Atta et al., 2023). As previously reported (Atta et al. 2023), maize individuals grown in the presence of HMs showed a significant reduction in aerial traits such as stem diameter, size of leaves, size and number of roots, and size of female inflorescences (Figure 1; Tables 1 and S1). On average, stem diameter showed a significant reduction of about 17% (P<0.05) as compared to plants grown in absence of HM stress (Figure 1f to 1j). Similarly, the length and width of developing leaves showed significant reductions of 12%, and mature leaves of 6% (P<0.05). Reproductive traits such as the number of female inflorescences (NFI) and their average weight (WFI) showed reductions, as previously reported (NFI = 1.20 ± 0.42 in absence of HMs and NFI = 0.70 ± 0.67 in presence of HMs; WFI = 40.06 ± 14.39 in absence of HMs and WFI = 37.75 ± 18.83 in presence of HMs; Table 1). However, the most significant effects occurred in the root system, with an increase in the number of brace roots under HMs (NBR = 12.90 ± 4.33 in absence of HMs; NBR = 16.30 ± 4.99 in presence of HMs; Table 2), and a significant decrease in their length (LBR = 16.15 ± 9.75 cm in absence of HMs; LBR = 4.00 ± 2.04 cm in presence of HMs; P<0.001; Table 1). The number and length of crown roots were reduced on average by 33% and 18%, respectively (NCR = 11.30 ± 2.11 in absence of HMs; NCR = 7.60 ± 0.97 in presence of HMs; LCR = 58.90 ± 21.87 cm in absence of HMs; LCR = 48.85 ± 11.80 cm in presence of HMs; Table 1). Nodal roots also showed a significant reduction in their number and length (P<0.01 and P<0.001, respectively; NNR = 10.50 ± 2.27 in absence of HMs; NNR = 6.70 ± 2.79 in presence of HMs; LNR = 87.20 ± 16.80 cm in absence of HMs; LNR = 55.30 ± 21.10 cm in presence of HMs; Table 1). By contrast, the number (NSR) and length (LSR) of seminal roots was significantly increased in the presence of HMs (P<0.001 and P<0.01, respectively: NSR = 4.10 ± 1.45 in absence of HMs; NSR = 6.60 ± 1.35 in presence of HMs; LSR = 27.70 ± 13.38 cm in absence of HMs; LSR = 43.50 ± 14.01 cm in presence of HMs) when measured 50 days after transplant (Table 1).

Phenotypic response of teosinte parviglumis and wild-type W22 maize grown under heavy metal stress.

(a) to (e), teosinte parviglumis; (f) to (j), W22 maize. (a) Teosinte parviglumis in absence of heavy metal stress at early flowering stage. (b) Teosinte parviglumis grown under heavy metal stress at early flowering stage. (c) Teosinte parviglumis in absence of heavy metal stress at late flowering stage. (d) Teosinte parviglumis grown under heavy metal stress at late flowering stage. (e) Teosinte parviglumis showing absence of lateral branching, and proliferation of female inflorescences arising from independent axillary meristems under heavy metal stress (detail of d). (f) W22 maize in absence of heavy metal stress at early flowering stage. (g) W22 maize grown under heavy metal stress at early flowering stage. (h) W22 maize in absence of heavy metal stress at late flowering stage. (i) W22 maize grown under heavy metal stress at late flowering stage. (j) W22 maize female inflorescences at late flowering stage (detail of i). Scale bars: 10 cm.

Interestingly, teosinte parviglumis exhibited a maize-like phenotype when grown in the presence of HMs, with complete absence of tillering in most plants (Figure 1; Tables 2, S1 to S3). This phenotype had never been reported for teosinte parviglumis plants grown under HM stress or any other type of edaphological conditions. From a total of 30 plants tested, 28 showed total absence of tillering, and two initiated secondary stems during vegetative stages but arrested their growth early during development, resulting in the presence of a single stem at reproductive stages (Figure 1a to 1d; and Table 2). Additionally, all 30 plants tested showed complete absence or residual lateral branching with unusual proliferation of female inflorescences arising from independent axillary meristems in primary stems (Figure 1e). This additional phenotype had not been reported for teosinte parviglumis under any type of environmental conditions. The increase in the number of female inflorescences clustered within a single stem (NFI = 1.80 ± 0.63 in absence of HMs; NFI = 3.90 ± 1.66 in presence of HMs; Figure 1e and Table 2) was associated with shorter flowering time as compared with normal growing conditions in absence of HMs (Figure 1d). In addition, teosinte parviglumis grown in the presence of HMs showed reduced leaf length (LIL = 71.30 ± 11.32 cm in absence of HMs; 56.25 ± 7.80 cm in presence of HMs; Table 2). The absence of lateral branches resulted in lower values of aerial dry weight (DWAP = 133.49 ± 28.28 gr in the absence of HMs; DWAP = 82.72 ± 11.67 gr in presence of HMs; Table 2). Teosinte parviglumis grown under HMs showed a significant reduction in the length of brace roots (P<0.01; LBR = 20.00 ± 11.49 cm in absence of HMs; LBR = 2.88 ± 0.66 cm in presence of HMs; Table 2), but no significant changes in their number. The number crown and nodal roots was significantly decreased by HM stress (P<0.001 and P<0.01 respectively; NCR=12.88 ± 4.34 in absence of HMs; 9.6 ± 1.43 in presence of HMs; NNR=16.2 ± 4.69 in absence of HMs; NNR= 14 ± 3.2 in presence of HMs; Table 2). By contrast, the length of nodal roots was significantly increased by HM stress (P<0.001; LNR=83 ± 15.76 in absence of HMs; 104 ± 8.46 in presence of HMs). The presence of HMs did not modify the absence of seminal roots intrinsic to teosinte parviglumis accessions (Perkins and Lynch 2021). Taken together, these results suggest that the response of maize and teosinte parviglumis to HM stress can give rise to phenotypes previously shown to distinguish both subspecies at the light of the domestication syndrome.

Genomic regions encompassing ZmHMA1, ZmHMA7 and ZmSKUs5 show reduced nucleotide variability in extant maize as compared to teosinte parviglumis

ZmHMA1 and ZmSKU5 mapped 52.1 and 41.8 cM, respectively, in a genetic map of chr.5 calculated to be 86.64 cM, with an average megabase pair to centimorgan ratio of 1.873 Mbp/cM (Lemmon and Doebley, 2014; Figure S1). The corresponding mapping population used isogenic recombinant inbred lines (NIRILs) with informative cross-over events concentrated in the region of interest and replicated phenotypic assessments in block experiments. Multiple QTLs were detected in a region spanning across 82 cM, none with single large effect genes such as those found in other chromosomes (Wang et al., 2005; Studer et al. 2011; Wills et al., 2013), suggesting that several associated domestication traits are due to multiple linked QTLs of small effect (Lemmon and Doebley, 2014). The location of ZmSKU5 overlaps with a single wide 1.5-LOD support interval that spreads across 50.6 cM (number of kernels per rank; Figure S1). By contrast, ZmHMA1 overlaps with a contiguous region containing no less than five QTL 1.5- LOD support intervals (ear diameter; number of kernels per rank; percentage of staminate spikelets; tassel branch number; and tillering index; Figure S1).

To determine if the regulatory and coding regions of ZmHMA1, ZmHMA7, and ZmSKUs5 were affected by human selection, we calculated their level of nucleotide variability based on single nucleotide polymorphisms (SNPs) for all extant maize and teosinte parviglumis accessions included in the diversity panel HapMap3, using the corresponding sequences of Tripsacum dactyloides as an outgroup. The levels of the nucleotide variability index - π(m) and π(t) for maize and teosinte parviglumis, respectively - were determined in bins of 100 nt using VCFtools, including alignments encompassing a genomic region that included 15 kb upstream of the START codon, the corresponding coding region for each gene, and 15 kb downstream of the STOP codon (Figure 2 and Table S4). In the case of ZmHMA1, the upstream region values were π(m)= 0.00071 and π(t)= 0.00870, and for the downstream region, π(m)=0.00097 and π(t)=0.00248. By contrast, the ZmHMA1 coding sequence showed low values in both subspecies: π(m)= 0.0050 and π(t)= 0.00106. These results indicate that human selection in the ZmHMA1 locus preferentially affected its upstream and downstream regulatory regions (Figure 2a and Table S4). ZmHMA7 showed contrasting values of π(m) and π(t) across the entire coding sequence and its regulatory regions (Figure 2b). For the upstream region, values were π(m)= 0.0009 and π(t)=0.00411; for the ZmHMA7 coding region, π(m)= 0.00015 and π(t)= 0.00148; and for the downstream region, π(m)=0.00066 and π(t)=0.00454. ZmSKUs5 also showed contrasting nucleotide variability in the entire coding and flanking regions (Figure 2c and Table S4). Values were π(m)= 0.0011 and π(t)= 0.00735 for the upstream region, π(m)= 0.00044 and π(t)= 0.00754 for the coding sequence, and π(m)=0.00203 and π(t)=0.00626 for the downstream region (Table S4). For all three genes, similar π values in the coding region of maize landraces and improved lines indicates that the loss of nucleotide variability occurred during the teosinte parviglumis to maize transition, and not during subsequent contemporary breeding (Table S4). By contrast, the ZmGLB1 neutral control locus shows equivalent levels of nucleotide variability in maize and teosinte parviglumis (Eyre-Walker et al. 1998).

Genetic diversity of loci encompassing ZmHMA1, ZmHMA7, and ZmSKUs5.

The nucleotide variability index (π) was calculated for all HapMap3 accessions of teosinte parviglumis (blue line), maize (red line), and Tripsacum dactyloides (orange line), taking in consideration the coding sequence (arrow) and an upstream and downstream region of 15 Kb encompassing each gene. (a) Nucleotide variability in the ZmHMA1 locus. (b) Nucleotide variability in the ZmHMA7 locus. (c) Nucleotide variability in the ZmSKUs5 locus. (d) Nucleotide variability in the ZmGLB1 locus.

We also compared the level of genetic variability in ZmHMA1, ZmSKUs5, and ZmHMA7 (and their neighboring regions) to all chr.5 genes categorized as likely affected by domestication (Hufford et al., 2012), and to the full collection of neutral genes distributed across the whole chromosome. The analysis was conducted using 100 bp bins and sliding windows of 50 bp; for a total of 385,829 bins in the case of chr.5 domestication genes, and 2,373,010 bins for neutral genes. As shown in Figure S2, the levels of nucleotide variability in the three candidate genes and their six neighboring loci are significantly reduced in maize landraces as compared to teosinte parviglumis inbred lines. When compared to the full neutral gene set of chr.5, this reduction is more severe than the reduction affecting 304 previously identified domestication genes, confirming the exceptional selective pressure that was imposed in those three loci during the evolutionary transition from teosinte parviglumis to extant maize.

Large-scale genomic and transcriptomic comparisons indicate that many HM response genes were positively selected across the maize genome

To expand the results well beyond the analysis of the three genes previously described, we performed a detailed analysis of genetic diversity across the 11.47 Mb genomic region comprised between ZmSKUs5 and ZmHMA1. This additional analysis reveals general tendencies in the quantity and nature of loci that were affected by positive selection during the teosinte parviglumis to maize transition in region identified via LOD score on chr.5. We compared nucleotide variability by using 100 bp bins covering loci composed of two 30 Kb segments up and downstream of coding sequences, respectively, and the coding sequence itself, for 173 genes present within the genomic region comprised between ZmSKUs5 and ZmHMA (Figure S1 and Supplementary File 6). Two types of statistical tests (ANOVA and Wilcoxon) were applied to nucleotide variability comparisons using the entirety of each locus. The Benjamini-Hochber procedure allowed an estimation of the false discovery rate (FDR<0.05) to avoid type I errors (false positives). Although some individual loci appear as differently classified depending on the statistical test applied (22 out of 173 loci), the general differences in nucleotide variability are consistently maintained within the subregions described below. We found that 166 out of 173 loci show signatures of positive selection and are roughly organized in five independent subregions of variable length. The first six loci are consecutively ordered in a 402 Kb subregion that includes ZmSKUs5. A second group of 13 consecutive loci expands over a 1.44 Mb subregion that contains NRAMP ALUMINUM TRANSPORTER1, also involved in HM response through uptake of divalent ions. A third group of 17 consecutive loci expands over 1.28 Mb; eleven contain genes encoding for uncharacterized proteins. The fourth group is composed of 57 consecutive loci expanding over 3.22 Mb and contains genes encoding for DEFECTIVE KERNEL55, AUXIN RESPONSE FACTOR16, and peroxydases involved in responses to oxydative stress. The fifth group contains 12 consecutive loci expanding over 713 Kb and contains ZmHMA1. An additional segment of approximately 1.17 Mb and containing 25 consecutive loci that were positively selected expands away from the ZmSKUs5-ZmHMA1 segment; it also contains several genes encoding for peroxydases. Although multiple loci include genes that could be involved in abiotic stress and oxidative responses, these results suggest that multiple factors other than HM stress could have played a role in the evolutionary mechanisms that affected the genetic diversity of chr.5 during the teosinte parviglumis to maize transition.

To further analyze the possibility that HM response could have played a role in maize emergence and subsequent domestication, we analyzed large scale transcriptomic data corresponding to independent experiments aiming at understanding the response of maize roots to HM stress. Six available transcriptomes were selected for in-depth analysis because they presented a fold change strictly higher than 1, and their results were supported by false discovery rates (FDR<0.05). These six transcriptomes (Table S5) included HM response datasets corresponding to growth conditions that not only incorporated Cu, but also lead (Pb) and chromium (Cr) that were not included in the substrate of our experiments. Transcriptional profiles were obtained from roots of plants at different stages: maize seedlings (Shen et al., 2012; Gao et al., 2015; Zhang et al., 2024a), three week old plantlets (Yang et al., 2023), and plants at V2 stage (Zhang et al., 2024b; Fengxia et al., 2025). A total of 120 genes shared by all six transcriptomes were found to be differentially expressed under HM stress conditions (66 upegulated and 54 downregulated; Figure S3), including ZmSKUs5, ZmHMA1 and ZmHMA7; 52 of them (43.3%) are located in maize loci showing less than 70% of the nucleotide variability found in teosinte parviglumis, suggesting that they were affected by positive selection (Yamasaki et al., 2005; Supplementary File 7). Of 18 mapping in chr.5, twelve are within the 82 cM that fractionates into multiple QTLs under selection during the parviglumis to maize transition. Interestingly, five additional loci containing HM response genes completely lack SNPs within their total length in both parviglumis and maize, and 19 additional loci lack SNPs in at least one 30 Kb segment or their coding region (Supplementary File 7), suggesting the frequent presence of ultraconserved genomic regions in many loci containing HM response genes. When this same analysis was conducted in a set of loci comprising 63 genes previously identified as differentially expressed in response to abiotic stress not directly related to HM responses (hypoxia; nutritional deficiency; soil alkalinity; drought; soil salinity), 18 loci (28.6%) showed less than 70% of the nucleotide variability found in teosinte parviglumis. Only one of them maps in chr.5 and none contained segments or coding regions lacking SNPs in parviglumis or maize. These results suggest that in contrast to other types of abiotic stress response genes, loci comprising a large set of genes that unambiguously respond to HM stress caused by chemical elements of diverse nature were affected by positive selection during the parviglumis to maize transition, irrespectively of their position in the genome.

ZmHMA1 and ZmHMA7 are active throughout development and their expression is influenced by the presence of heavy metals

The gene structure of ZmHMA1, ZmHMA7, and ZmSKUs5 is shown in Figure 3. The coding sequence of ZmHMA1 and ZmHMA7 is approximately 21.6 Kb and 4.5 Kb, respectively, while ZmSKUs5 is approximately 4.2 Kb in length. Within the short arm of chromosome 5, ZmHMA1 and ZmSKUs5 are separated by 9.16 Mb, with ZmHMA1 approximately 15 Mb away from the maximum likelihood prediction of a QTL involved in maize domestication (Doebley and Stec 1993; Vielle-Calzada et al. 2009). ZmHMA7 maps to the centromeric region of chromosome 5. A detailed PlantCARE-based analysis identified several conserved binding sites and regulatory elements in the promoter region of these three genes (Lescot et al. 2002). ZmHMA1 and ZmHMA7 share binding sites to basic Helix-Loop-Helix (bHLH) transcription factors, as well as abscisic acid response (ABRE), ethylene responsive (ERF), and G-box (CACGTG) elements recognized by leucine zipper family (bZIP) and bHLH transcription factors (Figure 3a). The regulatory region of both ZmHMA1 and ZmSKUs5 contain a binding site for MIKC-type MADS box transcription factors (Figure 3a), whereas ZmHMA7 and ZmSKUs5 share a jasmonic acid-response TGACG-motif (Figure 3a), as well as a Myb-binding (MBS) and GATA binding site. In addition, ZmHMA1 contains a binding site for TCP proteins such as Tb1, increasing the number of families of transcription factors that could be involved in shared or individual regulation of these three genes (Figure 3a).

Structure and gene expression of ZmHMA1 and ZmHMA7 in teosinte parviglumis and W22 maize.

(a) Structural analysis of ZmHMA1, ZmHMA7, and ZMSKUs5. Blue triangles indicate the location of motifs predicted in the regulatory region of each gene. The insertion site of Mu line LANMu1029790 is indicated by a red triangle. (b) RT-PCR analysis of ZmHMA1 and ZMHMA7 expression throughout development of teosinte parviglumis and W22 maize in presence or absence of heavy metal stress; L=leaf; S=stem; PR=primary root; NR=nodal root; CR=crown root. V2: 15 days after transplant; V6: 30 days after transplant; R1: 50 days after transplant. Controls: (-) absence of DNA; (+) genomic DNA. (c) qPCR analysis of ZmHMA1 expression in 1-month old primary roots, in absence or presence of heavy metal stress. (d) qPCR analysis of ZmHMA7 expression in 1-month old primary roots, in absence or presence of heavy metal stress. Histograms in (c) and (d) show the relative fold-change (normalized to actin, based on technical and biological triplicates); standard deviations are shown by vertical lines and Student T-test P-values represented by asterisks. * P<0.01; ** P<0.001.

To elucidate the expression patterns of ZmHMA1 and ZmHMA7, we conducted reverse transcriptase PCR (RT-PCR) in leaves, stems, roots, and flowers at two vegetative (V2 and V6) and one reproductive (R1) developmental stage, in teosinte parviglumis and wild-type W22 maize individuals (Figure 3b). In the case of maize, ZmHMA1 was expressed in leaves, stems, roots and flowers throughout development in either absence or presence of HMs. ZmHMA7 was also expressed in aerial organs throughout development in both absence and presence of HMs but was not expressed in roots at reproductive stage R1. In teosinte parviglumis, ZmHMA1 was also expressed throughout development in either absence or presence of HMs. By contrast, ZmHMA7 was constitutively expressed in aerial organs, but only expressed in roots under HM stress during V2 and V6 stages. To further determine if the level of expression of these two genes is influenced by the presence of HMs, we conducted quantitative real-time PCR (qPCR) using the primary root of teosinte parviglumis and maize plantlets at four weeks after germination (Figures 3c and 3d). In wild-type W22 maize, the expression of ZmHMA1 was not significantly different in the presence and absence of HMs. By contrast, in teosinte parviglumis the expression of ZMHMA1 was significantly increased in presence of HMs (Figure 3c). In the case of ZmHMA7, expression was significantly increased in both maize and teosinte parviglumis when plants were grown in the presence of HMs (Figure 3d). Overall, these results indicate that the activity of both ZmHMA1 and ZmHMA7 is influenced by HM stress in teosinte parviglumis and maize.

ZmHMA1 null mutant individuals show phenotypic defects related to the domestication syndrome

Transposon Mu insertional lines have been widely used to elucidate gene function in maize. To elucidate the function of ZmHMA1, we phenotypically analyzed LANMu1029790, a 4-generation backcross of line mu1029790 into the W22 background selected for single copy insertions in ZmHMA1 (McCarty et al. 2005). In LANMu1029790, a single Mu insertion is in the first exon of ZmHMA1, 65 bp downstream of the START codon (Figure 3a). RT-PCR analysis showed that expression of ZmHMA1 is completely absent in homozygous individuals for the LANMu1029790 insertion (Figure S4), confirming that the corresponding mutant is a null allele. Plants harboring a homozygous LANMu1029790 insertion are subsequently designated as zmhma1-m1::MuDR (abbreviated zmhma1-m1).

To determine the function of ZmHMA1 in maize, we compared the phenotype of wild-type W22 and homozygous zmhma1-m1 individuals at the reproductive R1 stage (Table 1). Under normal growing conditions, homozygous zmhma1-m1 individuals generated significantly less leaves than wild-type (P<0.001; NL = 13.4 ± 0.84 in wild-type, and NL = 11.4 ± 0.84 in zmhma1-m1). Whereas the length of mutant leaves was smaller than wild-type (LML = 77.4 ± 6.64 cm in wild-type, and LML = 69.85 ± 9.18 cm in zmhma1-m1), their width was significantly larger (P<0.001; WML = 6.75 ± 1.3 cm in wild-type, and WML = 9.75 ± 1.4 cm in zmhma1-m1), resulting in an increase of the total dry weight of aerial organs in mutant plants (DWAP = 18.43 ± 8.04 grams in wild-type, and DWAP = 28.46 ± 9.54 grams in zmhma1-m1). Homozygous zmhma1-m1 individuals also produced significantly more female inflorescences per plant (P<0.001; NFI = 1.2 ± 0.42 in wild-type, and NFI = 1.9 ± 0.32 in zmhma1) a difference that reflected in the total dry weight of female inflorescences per plant (P<0.001; WFI = 40.06 ± 14.39 gr in wild-type and WFI = 69.5 ± 11.2 gr in zmhma1); and significantly more spikes per male inflorescence (P<0.001; NSB = 9 ± 1.63 in wild-type, and NSB = 13.4 ± 2.32 in zmhma1-m1). Interestingly, the general length of roots was significantly smaller in homozygous zmhma1-m1 individuals, in particular nodal roots (P<0.001; LNR = 87.2 ± 16.8 cm in wild-type, and LNR = 50.9 ± 12.83 cm in zmhma1-m1). By contrast, mutant plants showed a significant increase in the number of seminal roots per plant (P<0.001; NSR = 4.1 ± 1.45 in wild-type, and NSR = 7.3 ± 1.16 in zmhma1-m1). When wild-type and mutant individuals were grown under HM stress, differences in the size of leaves, number of female inflorescences per plant, and number of spikes per male inflorescence were maintained. In addition, mutant individuals exposed to HM stress were significantly taller than wild-type (P<0.01; HEI = 82.2 ± 34.94 cm in wild-type, and HEI = 125.9 ± 18.81 cm in zmhma1-m1). In addition, total root and seminal root length were equivalent in wild-type and mutant individuals, mainly because the length of wild-type roots was significantly reduced under HM stress; however, the number of seminal roots was again significantly higher in mutant individuals (P<0.05; NSR = 6.6 ± 1.35 in wild-type, and NSR = 8.2 ± 1.62 in zmhma1-m1). These results indicate that in maize aerial organs, ZmHMA1 is involved in promoting leaf growth, restricting the number of female inflorescences per plant, and restricting the number of spikes in male inflorescences. In underground organs, ZmHMA1 is involved in promoting root growth and restricting the number of seminal roots per individual. Under HM stress, ZmHMA1 shows an additional role in restricting plant height, a function that was not identified under normal growing conditions. Overall, these results suggest that ZmHMA1 is involved in the control of some of the aerial and underground phenotypic traits that distinguish maize from teosinte parviglumis, especially under HM stress conditions.

Under heavy metal stress, Tb1 is overexpressed in the apical meristem of teosinte parviglumis

The absence of secondary ramifications in teosinte parviglumis grown under HM stress is reminiscent of the aerial architecture of wild-type maize. Previous QTL and molecular analysis suggested that Tb1 contributed to the architectural difference between maize and teosinte, as tb1 maize mutants develop secondary stem ramifications ending in male inflorescences, resembling teosinte in their overall architecture (Doebley et al. 1997). Interestingly, the regulatory region of ZmHMA1 contains a specific binding site for members of the Teosinte-branched 1/Cycloidea/Proliferating (TCP) family of transcription factors that includes Tb1 (Figure 3a). To investigate the possibility that the phenotypic response of teosinte parviglumis under HM stress could be regulated by Tb1, we conducted quantitative real-time PCR (qPCR) to determine the expression of ZmHMA1 and Tb1 in the shoot apical meristem (SAM) of one month old W22 maize and teosinte parviglumis plantlets (Figure 4a and 4b). The expression ZmHMA1 was downregulated in W22 wild-type maize meristems in the presence of HMs. By contrast, HM stress resulted in a significant increase of ZmHMA1 expression in teosinte parviglumis, confirming that in the maize ancestor the level of ZmHMA1 activity is influenced by the presence of HMs in the soil. In addition, the expression of Tb1 was significantly upregulated in teosinte parviglumis meristems in the presence of HMs. These results do not allow to directly link the regulation of ZmHMA1 expression to the function of Tb1; however, they open an opportunity to further investigate the possibility that under HM stress, the formation of secondary ramifications in teosinte parviglumis could be repressed by transcription factors of the TCP family, including Tb1.

Tb1 and ZmHMA1 expression in the shoot apical meristem of plantlets grown under heavy metal stress.

(a) qPCR analysis of Tb1 expression in 1-month old teosinte parviglumis plantlets grown in absence or presence of heavy metal stress. (b) qPCR analysis of ZmHMA1 expression in 1-month old teosinte parviglumis and W22 maize plantlets in absence or presence of heavy metal stress. Histograms show the relative fold-change (normalized to actin, based on technical and biological triplicates); standard deviations are shown by vertical lines and Student T-test P-values represented by asterisks; * P<0.01.

Paleoenvironmental studies reveal periods of climatic instability in the presumed region of maize emergence during the early Holocene

It is well accepted that temperature fluctuations, volcanism and anthropogenic impact shaped the distribution and abundance of plant species in the Transmexican Volcanic Belt (TMVB) during the last 14,000 years (Torrescano-Valle et al. 2019). The TMVB has produced close to 8000 volcanic structures (Ferrari et al., 2011), transforming the relief multiple times, and causing hydrographic and soil changes that actively modified the distribution and composition of plant communities in Central Mexico. Detailed paleoenvironmental data for the Pleistocene and Holocene is available for several lacustrine zones located within the 50 to 100 km range of the region currently considered the cradle of maize domestication (Matzuoka et al. 2002; Figure 5a). In Lake Zirahuén (102°44′ W; 19°26′ N and approximately 2075 meters above sea level; index [i] in Figure 5a), pollen, microcharcoal and magnetic susceptibility analyses of two sedimentary sequences reveals three periods of major ecological change during the early and middle Holocene. Between 9500 and 9000 calibrated years before present (cal yr BP), pine forests seem to have been associated with summer insolation increases. A second peak of forest change occurred at around 8200 cal yr BP, coinciding with cold oscillations documented in the North Atlantic. Finally, events occurred between 7500 and 7100 cal yr BP shows an abrupt change in the plant community related to humid Holocene climates and a presumed volcanic event (Lozano-García et al., 2013). The environmental history of the central Balsas watershed has also been documented by pollen, charcoal, and sedimentary analysis conducted in three lakes and a swamp of the Iguala valley (Piperno et al. 2007). Paleoecological records of lake Ixtacyola (8°20N, 99°35W and approximately 720 meters above sea level; index [ii] in Figure 5a) and lake Ixtapa (8°21N, 99°26W) indicate that an important increase in temperature and precipitation occurred between 13000 and 10000 cal yr BP. The pollen record of Ixtacyola showed that members of the genus Zea were already part of the vegetation coverage by 12900 to 13000 cal yr BP, suggesting that some teosintes – likely including parviglumis - were commonly found at elevation areas where they do not presently occur. Lake Almoloya (also named Chignahuapan; 19°05N, 99°20E and approximately 2575 meters above sea level; index [iii] in Figure 5a) in the upper Lerma basin is only 20 Km from the crater of the Nevado de Toluca that is responsible for creating the late Pleistocene Upper Toluca Pumice layer over which the Lerma basin is deposited. Pollen records indicate the presence of Zea species by 11080 to 10780 cal yr BP. As for other locations, an important period of climatic instability prevailed between 11500 and 8500 cal yr BP (Ludlow-Wiechers et al., 2005). Humidity fluctuations occurred until 8000 cal yr BP, with a stable temperate climate between 8500 and 5000 cal yr BP. Although pollen and diatom studies are often difficult to interpret at a regional scale, the overall reported results suggest consistent periods of Zea plants present in periods of environmental and climatic instability that correlate with the history of volcanic activity during the early Holocene, as described in the next section.

Location and age of volcanic eruptions occurred during the Holocene in comparison to collection sites of teosinte parviglumis populations most closely related to maize.

(a) Black asterisks correspond to the location and accession numbers of collected teosinte parviglumis populations that are phylogenetically most closely allied with extant maize, following Matzuoka et al. 2002. The location of 21 volcanoes belonging to the Michoacan-Guanajuato Volcanic Field is indicated with yellow circles. The locations of the Nevado de Toluca and Popocatéptl craters are indicated with a red and brown circle, respectively. Orange squares indicate the location of paleoenvironmental studies covering from 14,000 cal yr BP to current era, as described in Lozano-García et al. 2013 (i), Piperno et al. 2007 (ii), and Ludlow-Wiechers et al., 2005 (iii). (b) Time versus age of volcanic eruptions occurred in the map area of the Transmexican Volcanic Belt during the Holocene; numbers correspond to volcanoes illustrated in (a). Adapted from Macías and Arce, 2019.

Temporal and geographical convergence between volcanic eruptions and maize emergence during the Holocene

Current evidence indicates that the emergence and domestication of maize initiated in Mesoamerica some time around 9,000 yr BP (Matsuoka et al. 2002). The current location of teosinte parviglumis populations that are phylogenetically most closely allied with maize are currently distributed in a region located between the Michoacan-Guanajuato Volcanic Field (MGVF) at their northwest, and the Nevado de Toluca and Popocatéptl volcanoes at their east and northeast (Figure 5a; Matsuoka et al. 2002). Precise records of field data indicate that ten accessions were collected in the Balsas river drainage near Teloloapan and Sierra de Huautla (Guerrero), at approximately 100 km south of the Nevado de Toluca crater (Matzuoka et al. 2002). Three other accessions were collected near Tejupilco de Hidalgo and Zacazonapan (Estado de México), at approximately 50 to 60 km from the Nevado de Toluca crater (8762, JSG y LOS-161, and JSG-391). And four other accessions were located in Michoacan, at a location within the MGVF (accession 8763), or at mid-distance between the MGVF and the Nevado de Toluca crater (accessions JSG y LOS-130, 8761, and 8766).

The most important source of HMs in ancient soils of Mesoamerica is TMBV-dependent volcanic activity through short- and long-term effects related to lava deposits, ores, hydrothermal flow, and ash (Torrescano-Valle et al. 2019). Interestingly, TMVB activity in the center of Mexico was particularly intense during the Holocene, as compared to other regions and time frames (Marcias and Arce, 2019). The Nevado de Toluca volcano produced one of the most powerful eruptions from central Mesoamerica in the Holocene, giving rise to the Upper Toluca Pumice deposit at 12621 to 12025 cal yr BP (Arce et al., 2003; Figure 5b). The pumice fallout blanketed the Lerma and Mexico basins with 40 cm of coarse ash (Bloomfield and Valastro 1977; Arce et al. 2003). A second eruption dated by 36Cl exposure occurred at 9700 cal yr BP (Arce et al. 2003; Figure 5b), and the most recent eruption occurred at 3580 to 3831 cal yr BP (Macías et al. 1997). During the early and middle Holocene, the Popocatéptl volcano produced at least four eruptions dated 13037-12060, 10775–9564, 8328-7591, and 6262-5318 cal yr BP (Siebe et al. 1997); three other important eruptions occurred during the late Holocene, between 2713 and 733 cal yr BP (Siebe and Macías, 2006). In addition, the MGFV is a monogenetic volcanic field for which 23 independent eruptions have been documented during the Holocene, 21 of them located towards the southern part of the field, in close proximity to the region harboring some of the teosinte parviglumis populations most closely related to maize. Three of these eruptions occurred in the early Holocene (El Huanillo 1130 to 9688 cal yr BP; La Taza 10649 to 10300 cal yr BP; Cerro Grande 10173 to 9502 cal yr BP; Figure 5b), and three others during the initial period of the middle Holocene, between 8400 and 7696 cal yr BP (La Mina, Los Caballos, and Cerro Amarillo; Figure 5b). On average, a new volcano forms every ∼435 years in the MGFV (Macías and Arce, 2019). No less than 16 other eruptions occurred between 7159 cal yr BP and the present time (Figure 5b). Soils of volcanic origin (andosols) are currently distributed in regions north-west from the Nevado de Toluca and Popocatéptl craters, in close proximity with teosinte parviglumis populations most closely related to maize (Figure S5). Although modern distribution of teosinte populations may differ from their distribution around 9000 yr BP, and unknown populations more closely related to maize may yet to be discovered, this data indicates that the date and region where maize emerged is convergent with the dates and locations of several volcanic eruptions occurred during the Holocene in that same region.

Discussion

Although the evolutionary transition from teosinte parviglumis to maize resulted in important phenotypic divergence between landraces and their wild progenitor, the relevance of teosinte parviglumis phenotypic plasticity and its genetic consequences during domestication remain poorly understood. Our study offers the first evidence showing that a specific environmental factor (HM stress) influenced the teosinte parviglumis to maize transition through specific gene function (HM-response ATPases). The comparison of HM responses between both subspecies indicates that teosinte parviglumis phenotypically responds to HM stress acquiring phenotypes reminiscent of maize. Low temperatures and low CO2 concentrations caused a significant decrease in lateral branching but not in tillering. This decrease of lateral branching correlated with a decrease in teosinte seed yield (Piperno et al. 2015). Similar results were reported for parviglumis plants grown under low Nitrogen conditions (Gaudin et al. 2011). Our results under HM stress also show an increase in the number of female inflorescences per plant. This HM stress-dependent proliferation of several closely associated female inflorescences in a single teosinte parviglumis stem has not been previously reported. Female inflorescences arose from individual axillary meristems in multiple internodes, significantly increasing teosinte parviglumis seed yield. Under HM stress, we also show that Tb1 is overexpressed in the apical meristem of teosinte parviglumis, suggesting that formation of secondary ramifications is repressed by Tb1 function under HM stress, as in extant maize. At this stage we cannot discard the possibility that Tb1 upregulation in parviglumis reflects a more generalized response to abiotic stress; however, the expression ZmHMA1 is downregulated in W22 wild-type maize meristems in the presence of HMs but upregulated in teosinte parviglumis meristems, suggesting that a specific regulatory shift relating HM responses and ZmHMA1 function occurred during the teosinte parviglumis to maize transition.

We also show that ZmHMA1, ZmHMA7 and ZmSKUs5 exhibit reduced genetic diversity in extant maize as compared to teosinte parviglumis, suggesting that all three loci were affected by the evolutionary transition that gave rise to maize. The maize regulatory region and downstream 3’ area of the ZmHMA1 locus exhibit pronounced reduction in SNPs compared to teosinte parviglumis lines, likely affecting the molecular scope of its regulatory proteins, including transcription factors of the TCP family (Martín-Trillo and Cubas 2010). By contrast, ZmHMA7 and ZmSKUs5 showed a selective sweep across the entire maize locus. Although loss of genetic diversity is usually the result of human selection during domestication, it can also represent a consequence of natural selective pressures favoring fitness of specific teosinte parviglumis allelic variants better adapted to environmental changes and subsequently affected by human selection during the domestication process. This possibility is reflected by widely spread selective sweeps affecting a large portion of chr.5 that contains hundreds of genes showing signatures of positive selection. The analysis of 11.47 Mb covering the ZmHMA1-ZmSKUs5 segment confirms the presence of large but discrete genomic subregions that were positively selected during the teosinte parviglumis to maize transition. Although several contain genes involved in HM response and oxidative stress, the diversity of gene functions does not necessarily favor abiotic stress over other factors that could be at the origin of selective forces affecting these regions. By contrast, a large scale transcriptomic survey indicates that genes consistently responding to HMs (Cu, Cd, Pb and Cr ) show signatures of positive selection at unusual high frequencies (43.3%) as compared to loci containing genes responding to other types of abiotic stress (28.6%). Our identification of HM response genes affected by positive selection is far from being exhaustive. Nevertheless, it agrees with the expected effects of a widespread selective sweep caused by environmental changes that influenced the parviglumis to maize transition at the genetic level. Of intriguing interest are 24 loci that partially or completely lack SNPs in both teosinte parviglumis and maize, suggesting possible genetic bottlenecks occurred before the teosinte to maize transition. Examples of other edaphological factors driving genetic divergence either in the teosintes or maize include local adaptation to phosphorus concentration in mexicana and parviglumis (Aguirre-Liguori et al. 2019), and fast maize adaptation to changing iron availability through the action of genes involved in its mobilization, uptake, and transport (Benke and Stich 2011). Our results reveal a teosinte parviglumis environmental plasticity that could be related to the function of HM response genes positively selected during the teosinte parviglumis to maize transition. Previous studies have demonstrated that transposable elements (TEs) contribute to activation of maize genes in response to abiotic stress, affecting up to 20% of the genes upregulated in response to abiotic stress, and as many as 33% of genes that are only expressed in response to stress (Makarevitch et al., 2015). It is therefore possible that the HM response of some specific genes that influenced maize emergence or domestication could be mediated by TEs influencing or driving their transcriptional regulation.

The maize genome contains 12 genes encoding for members of the P1b family of ATPases (Cao et al. 2019). They are subdivided according to their preferential substrate affinity: the Copper (Cu)/Silver (Ag) group includes ZmHMA1 to ZmHMA4; and the Zinc (Zn)/Cobalt (Co)/Cadmium (Cd)/Lead (Pb) includes ZmHMA5 to ZmHMA12. Initial studies suggested that ZmHMA1 is expressed at low levels in vegetative, reproductive or root organs throughout development, contrary to ZmHMA7 that is strongly expressed in developing seeds 10 days after pollination; however, both genes appeared downregulated in plants grown under either Cu or Cd stress (Zhiguo et al. 2018). Additional studies by Cao et al (2019) indicated low expression of ZmHMA1 and high levels of ZmHMA7 expression in 12 tissues: roots, nodes, shoots, leaves, tassel, shoot apical meristem and young stem, internodes, anthers, silks, whole seedlings, endosperm, embryo, and pericarp. Under Cd stress, expression of both ZmHMA1 and ZmHMA7 was significantly upregulated in roots and downregulated in stems (Cao et al. 2019), contradicting previous results by Zhiguo et al. (2018). In Arabidopsis and barley, the corresponding ZmHMA1 structural homologs (AtHMA1 and HvHMA1, respectively) can transport a wide range of metals into the chloroplast (Mikkelsen et al. 2012; Boutigny et al. 2014). In the case of the Arabidopsis structural homolog of ZmHMA7, AtHMA7 (also called RAN1) is proposed to act as a metallochaperone located in the Golgi apparatus, delivering Cu to the ethylene receptor located in the plasma membrane (Hirayama et al. 1999). Our evidence complements previous studies by showing that in maize, ZmHMA1 root expression is restricted to vegetative developmental stages, and that ZmHMA7 is expressed in roots in response to HM stress. We also show through qPCR analysis that under HM stress, the expression of both ZmHMA1 and ZmHMA7 is upregulated in teosinte parviglumis primary roots of 4-week-old plantlets. Overall, these results confirm that both ZmHMA1 and ZmHMA7 are HM response genes active throughout teosinte and maize development.

In maize, the phenotypic analysis of null mutant individuals indicates that ZmHMA1 is involved in the control of aerial and underground developmental traits, some of which are related to important functional adaptations that distinguish maize from its wild ancestor. In addition to changes in leaf number, leaf length, and number of spikes per male inflorescence, zmhma1-m1 mutant individuals grown under normal edaphological conditions showed an increase in the number of female inflorescences per plant, a reduction in the length of nodal roots, and an increase in the number of seminal roots per plant. Whereas the promotion of nodal root growth is a distinctive trait of maize phenotypic plasticity and adaptation, previous studies showed that contrary to maize, plants of teosinte parviglumis lack seminal roots. This absence is also characteristic of paleobotanical maize specimens dating more than 5,000 cal yr BP, confirming that the acquisition of seminal roots is an important trait acquired during domestication (Lopez-Valdivia et al. 2022). The presence of seminal roots distinguishes extant maize from the teosintes (Burton et al. 2013), suggesting a possible involvement in domestication (Perkins and Lynch 2021; Lopez-Valdivia et al. 2022). Our results suggest that the allelic variants of the ZmHMA1 locus selected during domestication could have contributed to promote seminal root formation in maize, optimizing their number in accordance to seed endosperm availability and carbohydrate reserves (Perkins and Lynch 2021; Lopez-Valdivia et al. 2022). Interestingly, ZmHMA1 plays an important role in contributing to restricting plant height in response to HM stress, a phenotype that is also an important distinction between teosinte parviglumis and maize. Taken together, these results indicate that ZmHMA1 function was involved in the phenotypic transition from teosinte parviglumis to maize through the influence of HM stress; however, a better definition of its role in the origin and domestication of maize will require a functional analysis of the ZmHMA1 teosinte parviglumis homolog, and a detailed molecular analysis of the corresponding allelic variants and their phenotypic effects in both subspecies.

Ideas and Speculation

Our results point to an evolutionary model in which environmental changes in the form of edaphological HM stress imposed natural selective forces that acted on native teosinte parviglumis populations, causing phenotypic responses and a severe genetic bottleneck (Figure S6). Adapted plants developed phenotypic traits such as shorter height and emergence of seminal roots, progressively giving raise to primitive maize individuals with reduced genetic variation in HM response genes such as ZmHMA1. The subsequent action of human selection caused a drastic reduction in allelic variation of HM response genes, and genetic fixation of the resulting domestication traits.

The history of eruptive activity in the TMVB during the early Holocene coincides with the temporal estimation of initial stages in the evolutionary transition that gave rise to maize. Intense and recurrent eruptions occurred 12,000 to 6,000 years ago near the region presumed to be the cradle of maize domestication (Matsuoka et al. 2002; Macias and Arce, 2019). Located less than 60 Km from the region that harbors some of the natural teosinte parviglumis populations most closely related to maize (Matsuoka et al. 2002), the Nevado de Toluca volcano underwent three major eruptions approximately 12621-12025, 9700, and 3580- 3831 yr BP (Marcias and Arce, 2019). The large 9700 yr BP eruption caused in the region a period of climatic instability that lasted until 8500 yr BP (Ludlow-Wiechers et al., 2005). Could volcanic eruptions during the early Holocene have modified soil composition and influence the initial evolutionary transition from teosinte parviglumis to primitive maize? The possibility of environmental cues influencing maize emergence was mentioned by Benz and Iltis (1992), but not in relation to volcanic activity. Deciphering the ancient DNA sequence of HM response genes such as ZmHMA1 in paleobotanical maize specimens found in Central Mexico could help determine if its genetic diversity was already reduced by 5,300 to 6,200 yr BP (Ramos-Madrigal et al., 2016 ; Vallebueno-Estrada et al., 2016; Benz, 2001). Our results open the possibility for environmental changes caused by volcanic activity acting as a driving force in the selection of loci involved in adaptive responses such as heavy metal tolerance.

Materials and Methods

Plant materials and growth conditions

Seeds from maize inbred line W22, teosinte parviglumis accession CIMMYTMA 29791, and insertional line LANMu1029790 were sterilized using a 1:1 distilled water and bleach solution, rinsed three times in water, and germinated in MS solid medium. One week-old seedings were transplanted into PVC tubes (length: 90 cm; diameter: 20 cm) or 20-liter pots filled with a pre-sterilized mix composed of 30% peat moss, 10% vermiculite, 10% perlite, and 50% ground litter, complemented with 15-9-12 (N-P-K) OsmocoteTM fertilizer.

For HM growth conditions, 16 mg/kg of cadmium chloride (CdCl2) and 400 mg/kg of copper sulfate (CuSO4) were added to the substrate before seedling transplant. All plants were grown under glasshouse conditions at 28°C during day and 24°C during night, with relative humidity comprised between 40 and 50%.

Phenotypic analysis

All plants were harvested at 15 days (V2 stage), 30 days (V6 stage), and 50 days (R1 stage) after transplant. No less than ten plants per biological replicate were measured for each trait and stage, with three biological replicates per experiment. The width and length of immature leaves was measured in the 5th leaf. The length and width of mature leaves in the 3rd leaf. The total number of leaves, total number of stem internodes, plant height, basal stem diameter, total number of stem ramifications, and total fresh and dry weight of the aerial biomass were estimated at the R1 stage. Estimated root traits included the number and length of brace roots (aerial roots), the number and length of crown roots (underground roots originating within 2 cm below the basal node), number and length of nodal roots (underground roots originating more than 2 cm below the basal node), the number and length of seminal roots and maximum length of the root system. Statistical analyses were conducted using RStudio, version 4.3.0 (Horton and Kleinman 2015). The data are presented as mean ± standard deviation and were tested by a paired Student’s T-Test. When the effects were significant according to T-Test, the treatments were compared with the Welch two sample T-Test at P < 0.05.

Genetic diversity analysis

The SNP collection of maize chromosome 5 was analyzed on the basis of the HapMap V3.2.1 dataset (Bukowski et al. 2018), selecting individuals from maize landraces (LR) (Table S6) and teosinte parviglumis inbred line accessions (Teo; Table S7). A sliding window diversity (π) was calculated using VCFtools v0.1.13 (Danecek et al. 2011) with a window size of 100 bp (--window-pi 100) and a sliding step of 50 bp (--window-pi-step 50). Regions under selection were extracted from a list of 304 genes with domestication-related selection signatures across the whole chromosome (Hufford et al. 2012; Table S8), for a total of 385,829 sliding windows (Table S9 and S10). A second set of neutral gene regions comprising 2,373,010 windows was also extracted across the entirety of chromosome 5. These two collections were compared to ZmHMA1 (GRMZM2G067853), ZmHMA7 (GRMZM2G029951), and ZmSKUs5 (GRMZM2G076985), encompassing 2,743 windows (Table S11 and S12). Additionally, we analyzed genetic diversity in the neighboring regions of these candidate genes, which included six genes (GRMZM2G701784, GRMZM2G014508, GRMZM2G367857, GRMZM2G375607, GRMZM2G374375, and GRMZM2G067883), totaling 3,088 windows. We employed a custom Perl script (Vallebueno GitHub repository) to calculate two diversity metrics: fold change (FC= π(LR)/π(Teo)) and the proportional diversity (FP=π(LR)/(π(LR) + π(Teo)). Data visualization was performed using violin plots generated in R v4.2.2.

For an arbitrary locus encompassing 30 Kbs up and downstream of the coding sequence of ZmMA1, ZmHMA7, and ZmSKUs5, the number of segregating sites (S), the number of unique sequences (haplotypes, h), and the nucleotide diversity index (θ) were estimated using DnaSP version 6.0 for all accessions included in HapMap3 (Rozas et al. 2017; Bukowski et al. 2018). DnaSP was also used to estimate Tajima’s D, Fu and Li’s D, and Fu and Li’s F indexes (Tajima 1989; Fu and Li 1993). The average proportion of pairwise nucleotide differences per nucleotide site (π) was calculated with VCFtools using bins of 100 bp and steps of 25 bp (Tajima 1983; Weir and Cockerham 1984). Null hypothesis of equality of parameters followed two-tailed tests as previously described (Rozas et al. 2017).

To characterize genetic diversity across a 11.47 Mb genomic interval encompassing the ZmHMA1 to ZmSKUs5 segment on chr.5, we quantified nucleotide diversity (π) differences between maize and teosinte parviglumis. For each gene found within the segment, three regions were defined: an upstream region, the coding sequence (CDS), and a downstream region. Nucleotide diversity for each subspecies was calculated within each region and expressed as the ratio πMaize/πTeosinte (πM/πT). Subregions in which maize retained less than 70% of teosinte parviglumis diversity (π_M/π_T ≈ 0.7) were highlighted in Supplemental File 6. Genome-wide π estimates were derived from non-overlapping 100 bp windows that had been intersected between species. Two complementary statistical approaches were used to evaluate species differences per region. First, a parametric analysis of variance (ANOVA) was fitted for each region using π as the response variable and subspecies (maize vs teosinte parviglumis) as a fixed factor. ANOVA was used when its assumptions (approximate normality of residuals and homogeneity of variances) were considered acceptable following Fischer. Second, a paired Wilcoxon signed-rank test was applied to the paired π values from maize and teosinte parviglumis within each region as a distribution-free alternative robust to departures from normality and to outliers (Wilcoxon, 1945). P-values from both methods were adjusted to control the false discovery rate (FDR) using the Benjamini–Hochberg procedure (Benjamini & Hochberg, 1995). All data processing and statistical analyses were performed in R (version 4.3.1). Gene annotations were obtained from Phytozome using the Zea mays PHJ40 v1.2 annotation (Bornowski et al., 2021).

Gene expression analysis

Leaves, stems, tassels, female inflorescences, as well as primary, nodal, and crown roots were manually isolated, frozen in liquid nitrogen and stored at 4°C. Total RNA was extracted using TRIZOL (Thermo Fisher Scientific, USA). For reverse-transcriptase PCR (RT-PCR), RNA was converted to coding DNA using the One-step RT-PCR kit (QIAGEN, USA). The full list of primers is presented in Table S8. ZmCDK was used as a positive control. PCR amplification was conducted using the following parameters: at 94°C for 5 min.; 94°C 30 sec., 60°C 30 sec., 72°C 1 min. for 35 cycles; final stop after 72°C for 10 min. For quantitative real-time PCR (qPCR), total RNA was isolated from 4-week-old meristems using TRIZOL. Complementary DNA (cDNA) was synthesized using 4 μg of total RNA, 10 mM oligodTs and Superscript reverse transcriptase II (Invitrogen, USA). Primers for PCR were designed using the online program Primer 3 Plus (v.0.4.0) and verified with Oligocalculator (Sigma Chemical, St Louis MO) to discard dimer formation. PCR efficiencies of target and reference gene were determined by generating standard curves based on serial dilutions prepared for cDNA templates. PCR efficiency was calculated according to the slope of the standard curve (primers with 100% efficiency, the fold equals 2). Each quantitative real-time PCR (qPCR) reaction was performed in a 10 μl volume consisting of 5 μl of 23 SYBR Green PCR reaction Mix (Applied Biosystems, Foster City, CA). 3.5 μl of the DNA template (100 ng/ml), 0.5 μl of the forward primer (5 mM), 0.5 μl of the reverse primer (5 mM), and 0.5 μl of ultrapure water. The qPCR reactions were performed using the CFX96 Touch Real-Time PCR detection System and the data were analyzed using the Bio-Rad CFX Manager software v3.1. The thermal profile consisted of 10 min at 95°C, 40 cycles of 15 sec at 95°C, and 1 min at 60°C. Amplification results were collected at the end of the extension step. Primer sequences for qPCR amplification are listed in Table S7. A comparative 2-ΔΔCT method was used for determining a relative target quantity of gene expression (Schefe et al. 2006), and the cyclin-dependent kinase gene ZmCDK (GRMZM2G149286) was used as a control (Hu et al. 2019). Reproducibility of the results was evaluated for each sample by running three technical and three biological replicates of each of the reactions and each sample.